---
title: 'Temp/Size Analysis'
author: "Kathleen Durkin"
date: "2023-10-12"
output:
  pdf_document: default
  html_document: default
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, 
                      fig.height=3, fig.width=8,fig.align = "center")

library(tidyverse)
library(infer)
library(broom)
```

I'm doing this analysis under the assumption that all three of the SL columns (SL_11212022, SL_12272022, and SL_mm) are measuring some consistent size/length value over three measurement dates. Similarly, I'm assuming the the columns WWT_11212022, WWT_12272022, and WholeBodyWW_g are measurements of body weight (maybe whole body wet weight?) on the same three measurments dates. 

# Data Munging
```{r}
# read in the data
codTempData <- read.csv("../data/temp-experiment.csv")
head(codTempData)

# Create two new columns indicating change from Nov.2022 measurement to Feb.2022 measurement, for both size (mm) and weight (g). Also, modify the Temperature variable from a numeric to a factor, since it's the treatment (will be necessary for ANOVA/TukeyHSD)
codTempData_plus <- transform(codTempData,
                              sizeChange_mm = SL_mm - SL_11212022,
                              weightChange_g = WholeBodyWW_g - WWT_11212022) %>%
  mutate(codTempData, Temperature = relevel(as.factor(Temperature), "0", "5", "9", "16"))
head(codTempData_plus)


# Reformatted data with single column for size values and single column for measurement values (and additional column indicating measurement date), enabling grouping by size/weight measurement date
# Not I renamed the final size and weight measurments to include the date 02/08/2023 -- this is just so that the measurement date column can have a consistent option despite final measurments happening between the days of 02/08/2023 and 02/10/2023.

codTempData_reformat <- codTempData_plus %>%
  rename(WWT_02082023=WholeBodyWW_g) %>%
  rename(SL_02082023=SL_mm) %>%
  pivot_longer(
    cols = c("SL_11212022", "SL_12272022", "SL_02082023", "WWT_11212022", "WWT_12272022", "WWT_02082023"),
    names_to = "var",
    values_to = "value"
  ) %>%
  separate(var, into = c("var", "date"), sep = "_") %>%
  pivot_wider(
    names_from = "var",
    values_from = "value"
  )
# Set the date variable to have desired chronological order
codTempData_reformat$date <- factor(codTempData_reformat$date, levels = c("11212022", "12272022", "02082023"))
  
#mutate(date = relevel(as.factor(date), "11212022", "12272022", "02082023"))
  
```

# Plots
```{r}
codTempData_reformat %>%
  ggplot(aes(x=Temperature,
             y=SL,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2,
              height = 0.2,
              size = 1.5) +
  xlab("Temperature Treatment (*C)") +
  ylab("Size (mm)") +
  facet_wrap(~date)

codTempData_reformat %>%
  ggplot(aes(x=Temperature,
             y=WWT,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2,
              height = 0.2,
              size = 1.5) +
  xlab("Temperature Treatment (*C)") +
  ylab("Weight (g)") +
  facet_wrap(~date)
```

# Size/Weight Change
```{r}
codTempData_plus %>%
  ggplot(aes(x=Temperature,
             y=sizeChange_mm,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2,
              height = 0.2,
              size = 1.5) +
  xlab("Temperature Treatment (*C)") +
  ylab("Size Change (mm)")

codTempData_plus %>%
  ggplot(aes(x=Temperature,
             y=weightChange_g,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2,
              height = 0.2,
              size = 1.5) +
  xlab("Temperature Treatment (*C)") +
  ylab("Weight Change (g)")
```
Looking at these plots visually, there seems to be a difference in change in both size and weight over time among the treatment temperatures. Let's test this statistically

# Check Assumptions
```{r}
# Check conditions for ANOVA

# Normality
  # Not perfect, but normalish enough that I feel comfortable using ANOVA
codTempData_plus %>%
  ggplot(aes(sample = sizeChange_mm)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~Temperature)

codTempData_plus %>%
  ggplot(aes(sample = weightChange_g)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~Temperature)

# Variance
  # For both sizeChange and weightChange, the largest SD is, at most, ~2x the smallest SD which is sufficiently similar to allow usage of ANOVA (which is fairly robust against heterogeneity of variance). May want to also test using randomization to be safe.
codTempData_plus %>%
  group_by(Temperature) %>%
  summarize(meanSizeChange = mean(sizeChange_mm),
            sdSizeChange = sd(sizeChange_mm),
            meanWeightChange = mean(weightChange_g),
            sdWeightChange = sd(weightChange_g))

# Assuming data are independent (part of experimental design)

```

# ANOVA

## Size Change
```{r}
# ANOVA
sizeANOVA <- aov(sizeChange_mm~Temperature, data=codTempData_plus)

tidySizeANOVA <- tidy(sizeANOVA)
tidySizeANOVA

# Calculate R^2 (how much of the variation in the data is explained by the treatment)
r_squared <- tidySizeANOVA$sumsq[1]/(tidySizeANOVA$sumsq[1]+tidySizeANOVA$sumsq[2])
r_squared
```
p = 1.79e-18 << 0.05, so there is a significant relationship between treatment (temperature) and size growth (change in size). R^2=0.422, indicating ~42% of variance in size change is explained by the temperature treatment.

```{r}
# Tukey HSD
sizeANOVA %>%
  TukeyHSD() %>%
  tidy() %>% 
  select(contrast, estimate, adj.p.value) %>% 
  arrange(adj.p.value)

codTempData_plus %>%
  ggplot(aes(x=Temperature,
             y=sizeChange_mm,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, 
              height = 0.2, 
              size = 1.5) +
  annotate(geom = "text", x = 1:4, y = 32, label = c("A","B","C","C")) +
  xlab("Temperature Treatment (*C)") +
  ylab("Size Change (mm) from 11/21/22 to 02/08/23")

```


## Weight Change
```{r}
# ANOVA
weightANOVA <- aov(weightChange_g~Temperature, data=codTempData_plus)

tidyWeightANOVA <- tidy(weightANOVA)
tidyWeightANOVA

# Calculate R^2 (how much of the variation in the data is explained by the treatment)
r_squared <- tidyWeightANOVA$sumsq[1]/(tidyWeightANOVA$sumsq[1]+tidyWeightANOVA$sumsq[2])
r_squared
```
p = 3.73e-15 << 0.05, so there is a significant relationship between treatment (temperature) and weight change. R^2=0.362, indicating ~36% of variance in weight change is explained by the temperature treatment.

```{r}
# Tukey HSD
weightANOVA %>%
  TukeyHSD() %>%
  tidy() %>% 
  select(contrast, estimate, adj.p.value) %>% 
  arrange(adj.p.value)

codTempData_plus %>%
  ggplot(aes(x=Temperature,
             y=weightChange_g,
             group=Temperature)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, 
              height = 0.2, 
              size = 1.5) +
  annotate(geom = "text", x = 1:4, y = 12, label = c("A","B","C","C")) +
  xlab("Temperature Treatment (*C)") +
  ylab("Weight Change (g) from 11/21/22 to 02/08/23")

```

For both size and weight, growth from 11/21/22 to 02/08/23 significantly differed among all temperature treatments, with the exception of the 9 degree and 16 degree treatments. For the 9 and 16 degree treatments, changes in size and weight were statistically similar.