---
title: "21-cod-integration-crosswalk"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Do not evaluate by default; run manually after checking paths
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Overview

This is the **first notebook in the integrated RNA-seq + DNA methylation series**
(`21`–`24`). It builds the **sample crosswalk / integrated metadata table** that links
the three sample-identifier systems used in this project so that downstream notebooks can
join expression and methylation features for the same individual fish.

The three identifier systems are:

| System | Example | Source |
|--------|---------|--------|
| RNA-seq sample ID | `sample_1` | `data/DESeq2_Sample_Information.csv` (`sample_name`) |
| featureCounts column / BAM label | `1` (or `19-G`, `RESUB-36`) | `output/06.2-cod-RNAseq-alignment-genome/featureCounts-gene/featureCounts_gene_matrix.txt` |
| WGBS methylation ID | `01B` | `code/17.samplesheet.csv` (`sample`) |

The **linking key is `GeneticSamplingCount`** (the integer assigned at dissection,
recorded in `data/temp-experiment.csv`):

- RNA-seq IDs are `sample_{GeneticSamplingCount}`.
- WGBS IDs are `{GeneticSamplingCount, zero-padded to 2 digits}B` (e.g. `1` -> `01B`).

This notebook fills the gap noted in the repo: **there is currently no committed file
linking methylation `{NN}B` IDs to RNA-seq `sample_{N}` IDs.** The outputs here are the
canonical metadata used by `22`, `23`, and `24`.

> **NOTE / TODO for the lab.** The `{NN}B` ↔ `sample_{N}` mapping below is derived
> purely from the *numeric component* of each ID (both assumed to equal
> `GeneticSamplingCount`). This convention is consistent with the documented sample
> naming, but it has **not been independently verified against dissection records**.
> Before trusting the paired analyses, confirm that, for each paired individual, the
> WGBS tube `{NN}B` and the RNA-seq tube `sample_{N}` came from the **same fish and the
> same tissue (Liver)**. See the "Validation" section at the end.

## Inputs

- `data/DESeq2_Sample_Information.csv` — RNA-seq sample sheet (`sample_name`,
  `sample_number`, `tank`, `temp_treatment`, `tissue_type`).
- `data/temp-experiment.csv` — full dissection/experiment table; contains
  `GeneticSamplingCount`, `Microchip ID`, `Tank`, `Temperature`, tissue masses, etc.
- `code/17.samplesheet.csv` — WGBS methylseq samplesheet (`sample`, `fastq_1`, `fastq_2`).
- `output/06.2-cod-RNAseq-alignment-genome/featureCounts-gene/featureCounts_gene_matrix.txt`
  — used only to read the *column labels* so we know exactly which RNA-seq libraries
  exist (including resequenced/odd labels like `19-G`, `RESUB-36`).

## Outputs (written to `output/21-cod-integration-crosswalk/`)

- `integrated-sample-metadata.csv` — one row per known sample (RNA-seq and/or WGBS).
- `paired-sample-metadata.csv` — paired subset (both assays present).
- `crosswalk-summary.csv` — counts by assay availability × `temp_treatment`.

# Load packages

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)
```

# Paths and configuration

Paths are resolved relative to the repository root. When knitting from `code/`, the
working directory is `code/`, so we step up one level. You can also set the environment
variable `PROJECT_COD_TEMP_ROOT` to point at the repo root explicitly.

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)
message(glue("Project root: {project_root}"))

deseq2_info_path   <- file.path(project_root, "data/DESeq2_Sample_Information.csv")
temp_exp_path      <- file.path(project_root, "data/temp-experiment.csv")
meth_samplesheet   <- file.path(project_root, "code/17.samplesheet.csv")
featurecounts_path <- file.path(
  project_root,
  "output/06.2-cod-RNAseq-alignment-genome/featureCounts-gene/featureCounts_gene_matrix.txt"
)

output_dir <- file.path(project_root, "output/21-cod-integration-crosswalk")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

# Helper: parse `GeneticSamplingCount` from each ID system

```{r helpers, eval=TRUE}
# RNA-seq "sample_12" -> 12 ; methylation "01B" -> 1 ; "RESUB-36" -> 36
parse_count_from_rnaseq <- function(x) {
  as.integer(str_extract(x, "(?<=sample_)\\d+"))
}

# WGBS "01B" -> 1. Strip the trailing "B" and any leading zeros.
parse_count_from_meth <- function(x) {
  as.integer(str_extract(x, "^\\d+"))
}

# featureCounts BAM labels embed the count, sometimes with suffixes:
#   "1", "100", "19-G", "19-S", "RESUB-36" ...
# Pull the FIRST integer run. Keep the raw label so odd libraries are visible.
parse_count_from_bamlabel <- function(x) {
  as.integer(str_extract(x, "\\d+"))
}

# Format a GeneticSamplingCount integer as the WGBS-style "{NN}B" ID (2-digit pad).
meth_id_from_count <- function(n) {
  ifelse(is.na(n), NA_character_, sprintf("%02dB", n))
}
```

# RNA-seq metadata

```{r load-rnaseq-meta}
deseq2_info <- readr::read_csv(deseq2_info_path, show_col_types = FALSE)
glimpse(deseq2_info)

# Standardize to the integrated schema columns we care about.
rnaseq_meta <- deseq2_info %>%
  transmute(
    rnaseq_sample        = sample_name,                 # e.g. "sample_1"
    genetic_sampling_count = parse_count_from_rnaseq(sample_name),
    tank                 = tank,
    temp_treatment       = temp_treatment,
    tissue_type          = tissue_type
  ) %>%
  mutate(has_rnaseq = TRUE)

rnaseq_meta
```

## Cross-check against the actual featureCounts columns

The featureCounts matrix is the authoritative list of RNA-seq libraries that were
actually quantified. We read just the header so we know which libraries exist (and can
flag any that are *not* represented in `DESeq2_Sample_Information.csv`, e.g. resequenced
`RESUB-*` or split `19-G`/`19-S` libraries).

```{r featurecounts-columns}
# featureCounts output: line 1 is the program/command comment, line 2 is the header.
fc_header <- readr::read_tsv(featurecounts_path, skip = 1, n_max = 0,
                             show_col_types = FALSE)

# Drop the fixed annotation columns; the rest are BAM paths.
fixed_cols <- c("Geneid", "Chr", "Start", "End", "Strand", "Length")
bam_cols <- setdiff(names(fc_header), fixed_cols)

# BAM label = basename without ".sorted.bam"
bam_labels <- bam_cols %>%
  basename() %>%
  str_remove("\\.sorted\\.bam$")

featurecounts_libs <- tibble(
  bam_label = bam_labels,
  genetic_sampling_count = parse_count_from_bamlabel(bam_labels)
)

featurecounts_libs

# TODO (lab): decide how to treat non-standard libraries when integrating, e.g.
#   - "19-G"/"19-S", "20-G"/"20-S", "57-G"/"57-S", "58-G"/"58-S": gonad/somatic splits?
#   - "RESUB-36", "RESUB-76", ...: resequenced libraries (use which one?).
# For the paired-sample integration these are unlikely to matter (paired set is small),
# but document the decision here so 22 reads the intended column for each individual.
```

# WGBS (methylation) metadata

```{r load-meth-meta}
meth_sheet <- readr::read_csv(meth_samplesheet, show_col_types = FALSE)
glimpse(meth_sheet)

meth_meta <- meth_sheet %>%
  transmute(
    meth_sample = sample,                               # e.g. "01B"
    genetic_sampling_count = parse_count_from_meth(sample)
  ) %>%
  mutate(has_methylation = TRUE)

meth_meta
```

# Experiment/dissection table (for richer covariates)

`data/temp-experiment.csv` is keyed by `GeneticSamplingCount` and adds covariates that
may be useful downstream (microchip ID, body/liver masses, dissection date). Note that
column names contain spaces, so we reference them with backticks / `janitor`-style
cleaning.

```{r load-temp-experiment}
temp_exp <- readr::read_csv(temp_exp_path, show_col_types = FALSE)
glimpse(temp_exp)

temp_exp_keyed <- temp_exp %>%
  rename(genetic_sampling_count = GeneticSamplingCount) %>%
  # keep a compact set of covariates; expand as needed
  transmute(
    genetic_sampling_count,
    microchip_id   = `Microchip ID`,
    exp_tank       = Tank,
    exp_temp       = Temperature,
    sl_mm          = SL_mm,
    whole_body_ww_g = WholeBodyWW_g,
    total_liver_ww_mg = TOTAL_Liver_WW_mg,
    dissection_date = DissectionDate
  ) %>%
  filter(!is.na(genetic_sampling_count)) %>%
  distinct(genetic_sampling_count, .keep_all = TRUE)

temp_exp_keyed
```

# Build the integrated crosswalk

We full-join RNA-seq and methylation metadata on `genetic_sampling_count`, then attach
the dissection covariates. `paired = has_rnaseq & has_methylation`.

```{r build-crosswalk}
crosswalk <- full_join(
  rnaseq_meta,
  meth_meta,
  by = "genetic_sampling_count"
) %>%
  mutate(
    has_rnaseq      = coalesce(has_rnaseq, FALSE),
    has_methylation = coalesce(has_methylation, FALSE),
    # Re-derive the expected methylation ID from the count for any RNA-only rows,
    # so the lab can see what the WGBS ID *would* be if that fish were sequenced.
    expected_meth_sample = meth_id_from_count(genetic_sampling_count),
    paired = has_rnaseq & has_methylation
  ) %>%
  left_join(temp_exp_keyed, by = "genetic_sampling_count") %>%
  # Prefer the RNA-seq sample sheet's temp_treatment; fall back to experiment table.
  mutate(temp_treatment = coalesce(temp_treatment, exp_temp)) %>%
  arrange(genetic_sampling_count)

crosswalk
```

```{r write-crosswalk}
readr::write_csv(crosswalk, file.path(output_dir, "integrated-sample-metadata.csv"))
```

# Paired subset

These are the individuals that carry **both** assays in-repo. The repo notes ~10 such
individuals (`GeneticSamplingCount`: 1, 2, 3, 13, 21, 31, 49, 60, 67, 68); this chunk
derives the set programmatically rather than hard-coding it.

```{r paired-subset}
paired <- crosswalk %>%
  filter(paired) %>%
  select(
    genetic_sampling_count,
    rnaseq_sample,
    meth_sample,
    temp_treatment,
    tissue_type,
    tank,
    microchip_id
  ) %>%
  arrange(genetic_sampling_count)

message(glue("Paired individuals (both assays): {nrow(paired)}"))
paired

readr::write_csv(paired, file.path(output_dir, "paired-sample-metadata.csv"))
```

```{r expected-paired-check, eval=TRUE}
# Sanity reference: the paired set documented in the repo.
expected_paired_counts <- c(1, 2, 3, 13, 21, 31, 49, 60, 67, 68)
message(glue(
  "Documented paired GeneticSamplingCounts: {paste(expected_paired_counts, collapse = ', ')}"
))
# After running the chunks above, compare:
# setdiff(expected_paired_counts, paired$genetic_sampling_count)  # documented but missing
# setdiff(paired$genetic_sampling_count, expected_paired_counts)  # newly paired
```

# Summary table

```{r summary}
crosswalk_summary <- crosswalk %>%
  mutate(
    assay = case_when(
      has_rnaseq & has_methylation ~ "both",
      has_rnaseq                   ~ "rnaseq_only",
      has_methylation              ~ "methylation_only",
      TRUE                         ~ "none"
    )
  ) %>%
  count(temp_treatment, assay) %>%
  pivot_wider(names_from = assay, values_from = n, values_fill = 0) %>%
  arrange(temp_treatment)

crosswalk_summary
readr::write_csv(crosswalk_summary, file.path(output_dir, "crosswalk-summary.csv"))
```

# Validation checks (run before trusting paired analyses)

```{r validation}
# 1. Every paired individual should resolve to exactly one RNA-seq and one WGBS ID.
dup_check <- crosswalk %>%
  filter(paired) %>%
  count(genetic_sampling_count) %>%
  filter(n > 1)
stopifnot(nrow(dup_check) == 0)

# 2. Tissue should be Liver for paired individuals (RNA-seq is mostly Liver).
#    WGBS tissue is NOT recorded in 17.samplesheet.csv -> see TODO below.
paired_tissue <- crosswalk %>% filter(paired) %>% count(tissue_type)
paired_tissue

# 3. Methylation IDs we expected vs. methylation IDs that actually exist.
missing_expected_meth <- crosswalk %>%
  filter(has_rnaseq, !has_methylation) %>%
  pull(expected_meth_sample)
length(missing_expected_meth)
```

> **TODO / assumptions for the lab to resolve:**
>
> 1. **Tissue match.** `code/17.samplesheet.csv` does not record tissue for the WGBS
>    libraries. RNA-seq is "mostly Liver". Confirm each paired WGBS library is **Liver**
>    (or record its true tissue) before interpreting expression–methylation correlations;
>    a tissue mismatch would confound everything downstream.
> 2. **ID-to-fish provenance.** Verify the `{NN}B` ↔ `sample_{N}` mapping against
>    dissection sheets / microchip IDs, not just the numeric convention used here.
> 3. **Non-standard RNA-seq libraries.** Decide which featureCounts column to use for any
>    individual with `RESUB-*` or `-G`/`-S` split libraries (none are in the documented
>    paired set, but record the rule in case the paired set grows).
> 4. **`temp_treatment` source of truth.** We prefer `DESeq2_Sample_Information.csv` and
>    fall back to `temp-experiment.csv$Temperature`; confirm these agree for paired fish.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
