---
title: "22-cod-integration-feature-matrices"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Heavy compute / off-repo data: run manually
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Second notebook in the integrated RNA-seq + DNA methylation series (`21`–`24`). It
prepares **matched, paired-sample feature matrices** so the integration in `23` operates
on the same set of genes and the same individuals from both assays.

It produces three feature matrices restricted to the **paired samples** identified in
`21`:

1. **Gene expression** — normalized from the genome-aligned featureCounts gene matrix
   (`06.2`). We use DESeq2 variance-stabilizing transformation (VST) as the primary
   normalization, with an edgeR logCPM alternative provided.
2. **Gene-body methylation** — per-gene weighted methylation fraction from the methylseq
   `coverage2cytosine` CpG reports, using gene-body coordinates from the genome GTF.
3. **Promoter methylation** — same, but over a TSS-flanking window (default ±1 kb).

The common feature key across all matrices is the **NCBI gene ID** (e.g. `LOC132464423`),
which is the `gene_id` used by featureCounts (`-g gene_id`) and present in `genomic.gtf`.

## Inputs

- `output/21-cod-integration-crosswalk/paired-sample-metadata.csv` (from notebook `21`).
- `output/06.2-cod-RNAseq-alignment-genome/featureCounts-gene/featureCounts_gene_matrix.txt`.
- Methylation `coverage2cytosine` coverage files. **Path note:** notebooks `18`/`20`
  reference both `data/17-nextflow/coverage2cytosine/coverage/` and
  `output/17-nextflow/coverage2cytosine/coverage/`. Set `cov_dir` below to wherever the
  `*.cov.gz` files live in your environment (klone / Owl mirror).
- Genome annotation `data/genomic.gtf` (NCBI GCF_031168955.1; downloaded off-repo in
  `06.2`, not committed).

## Outputs (written to `output/22-cod-integration-feature-matrices/`)

- `expression-vst-paired.csv` — VST expression, genes × paired samples.
- `methylation-genebody-paired.csv` — gene-body weighted methylation.
- `methylation-promoter-paired.csv` — promoter weighted methylation.
- `common-genes.txt` — genes present in both expression and (gene-body) methylation.
- `feature-matrix-summary.csv` — dimensions and filtering notes.

# Load packages

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)
library(vroom)

# Bioconductor packages used in heavy chunks (install once):
# BiocManager::install(c("DESeq2", "edgeR", "GenomicRanges", "GenomicFeatures",
#                        "txdbmaker", "rtracklayer"))
quietly_require <- function(pkg) {
  if (suppressWarnings(requireNamespace(pkg, quietly = TRUE))) {
    library(pkg, character.only = TRUE)
    return(TRUE)
  }
  FALSE
}
```

# Paths and configuration

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)

paired_meta_path   <- file.path(project_root,
  "output/21-cod-integration-crosswalk/paired-sample-metadata.csv")
featurecounts_path <- file.path(project_root,
  "output/06.2-cod-RNAseq-alignment-genome/featureCounts-gene/featureCounts_gene_matrix.txt")
gtf_path           <- file.path(project_root, "data/genomic.gtf")

# TODO (lab): point this at the coverage2cytosine *.cov.gz directory in your env.
cov_dir <- file.path(project_root, "data/17-nextflow/coverage2cytosine/coverage")

output_dir <- file.path(project_root, "output/22-cod-integration-feature-matrices")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Tunables
min_cpg_coverage   <- 5L     # min read depth to keep a CpG (per sample)
min_cpgs_per_gene  <- 3L     # min CpGs overlapping a feature to report methylation
promoter_upstream  <- 1000L  # bp upstream of TSS
promoter_downstream <- 200L  # bp downstream of TSS (into the gene)
```

```{r load-paired-meta, eval=TRUE}
paired <- readr::read_csv(paired_meta_path, show_col_types = FALSE)
paired
# Expected columns: genetic_sampling_count, rnaseq_sample, meth_sample,
#                   temp_treatment, tissue_type, tank, microchip_id
```

# Gene expression matrix (RNA-seq)

## Import featureCounts and map columns to paired samples

featureCounts columns are BAM paths whose basename encodes the
`GeneticSamplingCount` (e.g. `.../13.sorted.bam` -> count 13). We map those to the
paired `rnaseq_sample` IDs.

```{r import-featurecounts}
# Line 1 is the featureCounts command comment; the real header is line 2.
fc_raw <- readr::read_tsv(featurecounts_path, skip = 1, show_col_types = FALSE)

annot_cols <- c("Geneid", "Chr", "Start", "End", "Strand", "Length")
count_mat <- fc_raw %>%
  select(Geneid, all_of(setdiff(names(fc_raw), annot_cols))) %>%
  column_to_rownames("Geneid") %>%
  as.matrix()

# Clean BAM-path column names down to their label, then to GeneticSamplingCount.
bam_labels <- colnames(count_mat) %>% basename() %>% str_remove("\\.sorted\\.bam$")
bam_counts <- as.integer(str_extract(bam_labels, "\\d+"))

colnames(count_mat) <- bam_labels

# Keep only the (standard) columns whose count is in the paired set. If an individual
# has multiple libraries (e.g. RESUB-*), this selects the plain label first.
paired_counts <- paired$genetic_sampling_count
keep_idx <- which(bam_counts %in% paired_counts & bam_labels == as.character(bam_counts))
count_mat_paired <- count_mat[, keep_idx, drop = FALSE]

# Reorder/relabel columns to the rnaseq_sample IDs, ordered like `paired`.
label_to_sample <- paired %>%
  mutate(label = as.character(genetic_sampling_count)) %>%
  select(label, rnaseq_sample) %>%
  deframe()

colnames(count_mat_paired) <- label_to_sample[colnames(count_mat_paired)]
count_mat_paired <- count_mat_paired[, paired$rnaseq_sample[
  paired$rnaseq_sample %in% colnames(count_mat_paired)], drop = FALSE]

dim(count_mat_paired)
head(count_mat_paired)

# TODO (lab): if any paired individual is missing here, it means its standard featureCounts
# column was not found (maybe only a RESUB-* / split library exists). Resolve per the rule
# you recorded in notebook 21.
```

## Filter low-count genes and normalize with DESeq2 VST

```{r deseq2-vst}
has_deseq2 <- quietly_require("DESeq2")
stopifnot(has_deseq2)

# Light pre-filter: keep genes with a few counts across the (small) paired set.
keep_genes <- rowSums(count_mat_paired >= 10) >= 3
count_filt <- count_mat_paired[keep_genes, , drop = FALSE]
message(glue("Genes kept after pre-filter: {nrow(count_filt)} / {nrow(count_mat_paired)}"))

coldata <- paired %>%
  filter(rnaseq_sample %in% colnames(count_filt)) %>%
  mutate(temp_treatment = factor(temp_treatment)) %>%
  column_to_rownames("rnaseq_sample")
coldata <- coldata[colnames(count_filt), , drop = FALSE]

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = count_filt,
  colData   = coldata,
  design    = ~ 1                # design-agnostic transform; modeling happens in 23/DE notebooks
)

vsd <- DESeq2::vst(dds, blind = TRUE)
expr_vst <- SummarizedExperiment::assay(vsd)

# Column names -> rnaseq_sample IDs (already are). Write genes x samples.
expr_vst_df <- as_tibble(expr_vst, rownames = "gene_id")
readr::write_csv(expr_vst_df, file.path(output_dir, "expression-vst-paired.csv"))
dim(expr_vst_df)
```

## Alternative: edgeR logCPM (optional)

```{r edger-logcpm}
has_edger <- quietly_require("edgeR")
if (has_edger) {
  dge <- edgeR::DGEList(counts = count_mat_paired)
  keep <- edgeR::filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- edgeR::calcNormFactors(dge)
  logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 2)
  logcpm_df <- as_tibble(logcpm, rownames = "gene_id")
  readr::write_csv(logcpm_df, file.path(output_dir, "expression-logcpm-paired.csv"))
  dim(logcpm_df)
} else {
  message("edgeR not installed; skipping logCPM alternative.")
}
```

# Gene-level methylation matrix (WGBS)

We summarize CpG methylation to **gene-level features** by overlapping each sample's
CpGs with (a) gene-body intervals and (b) promoter intervals from the genome GTF, then
computing a **coverage-weighted methylation fraction** per gene:

\[
\text{methylation}_{gene} = \frac{\sum_{\text{CpG} \in gene} \text{count\_meth}}{\sum_{\text{CpG} \in gene} (\text{count\_meth} + \text{count\_unmeth})}
\]

This weighted approach is standard (e.g. methylKit/bsseq region summaries) and is more
robust than averaging per-CpG percentages.

## Build gene-body and promoter ranges from the GTF

```{r build-ranges}
has_gr  <- quietly_require("GenomicRanges")
has_gf  <- quietly_require("GenomicFeatures")
stopifnot(has_gr, has_gf)

# txdbmaker::makeTxDbFromGFF in recent Bioconductor; falls back to GenomicFeatures.
make_txdb <- function(gtf) {
  if (quietly_require("txdbmaker")) {
    txdbmaker::makeTxDbFromGFF(gtf, format = "gtf")
  } else {
    GenomicFeatures::makeTxDbFromGFF(gtf, format = "gtf")
  }
}

txdb <- make_txdb(gtf_path)

# Gene bodies keyed by gene_id (matches featureCounts -g gene_id).
gene_bodies <- GenomicFeatures::genes(txdb)

# Promoters: TSS-flanking windows.
gene_promoters <- GenomicRanges::promoters(
  gene_bodies,
  upstream   = promoter_upstream,
  downstream = promoter_downstream
)

length(gene_bodies)
length(gene_promoters)

# IMPORTANT: seqlevels (chromosome names) in the GTF must match the coverage files.
# Both come from GCF_031168955.1, so they should be NC_/NW_ RefSeq accessions
# (e.g. "NC_082382.1"). Verify before overlapping:
# head(GenomeInfoDb::seqlevels(gene_bodies))
```

## Read coverage files for paired methylation samples

```{r read-coverage}
# Match the paired WGBS IDs (e.g. "13B") to coverage files in cov_dir.
all_cov <- list.files(cov_dir, pattern = "\\.cov(\\.gz)?$", full.names = TRUE)
stopifnot(length(all_cov) > 0)

cov_label <- function(path) sub("\\..*$", "", basename(path))  # "01B.something.cov.gz" -> "01B"

cov_index <- tibble(path = all_cov, meth_sample = map_chr(all_cov, cov_label))
paired_cov <- cov_index %>% filter(meth_sample %in% paired$meth_sample)
paired_cov

# coverage2cytosine .cov columns: chr, start, end, perc_meth, count_meth, count_unmeth
read_cov <- function(path, min_coverage = min_cpg_coverage) {
  vroom::vroom(
    path, delim = "\t",
    col_names = c("chr", "start", "end", "perc_meth", "count_meth", "count_unmeth"),
    col_types = list(chr = "c", start = "i", end = "i",
                     perc_meth = "d", count_meth = "i", count_unmeth = "i"),
    progress = FALSE
  ) %>%
    mutate(depth = count_meth + count_unmeth) %>%
    filter(depth >= min_coverage)
}
```

## Overlap CpGs with features and compute weighted methylation

```{r summarize-methylation}
# Returns a named numeric vector: gene_id -> weighted methylation, for one sample.
summarize_sample <- function(cov_df, feature_gr, min_cpgs = min_cpgs_per_gene) {
  cpg_gr <- GenomicRanges::GRanges(
    seqnames = cov_df$chr,
    ranges   = IRanges::IRanges(start = cov_df$start, end = cov_df$start),
    count_meth   = cov_df$count_meth,
    count_unmeth = cov_df$count_unmeth
  )
  hits <- GenomicRanges::findOverlaps(cpg_gr, feature_gr, ignore.strand = TRUE)
  if (length(hits) == 0) return(tibble(gene_id = character(), meth = numeric()))

  gene_idx <- S4Vectors::subjectHits(hits)
  cpg_idx  <- S4Vectors::queryHits(hits)

  tibble(
    gene_id      = names(feature_gr)[gene_idx],
    count_meth   = cov_df$count_meth[cpg_idx],
    count_unmeth = cov_df$count_unmeth[cpg_idx]
  ) %>%
    group_by(gene_id) %>%
    summarise(
      n_cpg = n(),
      meth  = sum(count_meth) / sum(count_meth + count_unmeth),
      .groups = "drop"
    ) %>%
    filter(n_cpg >= min_cpgs) %>%
    select(gene_id, meth)
}

# Build a long table across all paired methylation samples for a given feature set.
build_meth_matrix <- function(feature_gr) {
  paired_cov %>%
    mutate(data = map(path, ~ summarize_sample(read_cov(.x), feature_gr))) %>%
    select(meth_sample, data) %>%
    unnest(data) %>%
    pivot_wider(names_from = meth_sample, values_from = meth)
}

meth_genebody <- build_meth_matrix(gene_bodies)
meth_promoter <- build_meth_matrix(gene_promoters)

dim(meth_genebody)
dim(meth_promoter)
```

## Relabel methylation columns to the integrated sample key and write

```{r write-methylation}
# Optionally rename meth columns from "{NN}B" to the matching rnaseq_sample, so that
# downstream joins use a single shared sample key. We keep BOTH styles documented.
meth_to_rnaseq <- paired %>% select(meth_sample, rnaseq_sample) %>% deframe()

relabel_meth <- function(df) {
  cols <- intersect(names(df), names(meth_to_rnaseq))
  rename_vec <- setNames(cols, unname(meth_to_rnaseq[cols]))
  df %>% rename(!!!rename_vec)
}

readr::write_csv(meth_genebody, file.path(output_dir, "methylation-genebody-paired.csv"))
readr::write_csv(meth_promoter, file.path(output_dir, "methylation-promoter-paired.csv"))

# Also write rnaseq-keyed versions for convenience in 23/24.
readr::write_csv(relabel_meth(meth_genebody),
                 file.path(output_dir, "methylation-genebody-paired-rnaseqkey.csv"))
readr::write_csv(relabel_meth(meth_promoter),
                 file.path(output_dir, "methylation-promoter-paired-rnaseqkey.csv"))
```

# Common gene set and summary

```{r common-genes}
expr_genes <- expr_vst_df$gene_id
meth_genes <- meth_genebody$gene_id
common_genes <- intersect(expr_genes, meth_genes)
message(glue("Common genes (expression ∩ gene-body methylation): {length(common_genes)}"))

writeLines(common_genes, file.path(output_dir, "common-genes.txt"))

feature_summary <- tibble(
  matrix = c("expression_vst", "methylation_genebody", "methylation_promoter",
             "common_genes"),
  n_features = c(length(expr_genes), nrow(meth_genebody), nrow(meth_promoter),
                 length(common_genes)),
  n_samples  = c(ncol(expr_vst_df) - 1, ncol(meth_genebody) - 1,
                 ncol(meth_promoter) - 1, NA_integer_),
  note = c(
    "DESeq2 VST, blind, pre-filtered rowSums(counts>=10)>=3",
    glue("weighted meth; min_depth={min_cpg_coverage}; min_cpgs={min_cpgs_per_gene}"),
    glue("promoter window -{promoter_upstream}/+{promoter_downstream} bp around TSS"),
    "expression ∩ gene-body methylation gene IDs"
  )
)
readr::write_csv(feature_summary, file.path(output_dir, "feature-matrix-summary.csv"))
feature_summary
```

# Notes, assumptions, and TODOs

- **Coverage directory.** Set `cov_dir` to the actual location of the `*.cov.gz`
  `coverage2cytosine` files (klone `/gscratch/...` or Owl mirror). Only paired samples
  are read here.
- **Seqlevel matching.** The GTF and coverage chromosome names must match (both RefSeq
  `NC_/NW_` accessions from GCF_031168955.1). If not, harmonize with
  `GenomeInfoDb::seqlevelsStyle()` before `findOverlaps`.
- **Feature definition.** Gene-body and promoter (±1 kb / +200 bp) are reasonable defaults.
  The lab may prefer CpG-island-, exon-, or first-intron-level summaries; the
  `summarize_sample()` helper accepts any `GRanges` feature set.
- **Tissue caveat carries over from `21`.** These matrices assume each paired WGBS library
  is the same tissue as the RNA-seq library (Liver). Confirm before interpreting.
- **No DE/DMR pre-selection here.** This notebook produces *full* normalized matrices;
  feature selection (variable genes, DEGs, DMR genes) happens in `23`.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
