---
title: "23-cod-integration-analysis"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,        # Run manually after 21 and 22 have produced their outputs
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Third notebook in the integrated RNA-seq + DNA methylation series (`21`–`24`). It performs
the **expression–methylation integration** on the paired samples, combining a transparent
per-gene correlation analysis with a joint multivariate view.

Three complementary analyses:

1. **Per-gene expression–methylation correlation.** For each gene present in both
   matrices, correlate VST expression against gene-body (and promoter) methylation across
   the paired individuals, with multiple-testing correction (BH FDR).
2. **Joint PCA.** Scale and concatenate the two omics blocks and run a single PCA to see
   whether samples separate by `temp_treatment` in the combined space.
3. **Multi-omics integration (optional).** A mixOmics **DIABLO** (or sPLS) block model
   that selects correlated expression + methylation features discriminating temperature
   treatments.

> **Statistical caveat (read first).** The paired set is small (~10 individuals). Per-gene
> correlations across n≈10 have low power and wide CIs; treat gene-level results as
> hypothesis-generating and lean on the multivariate (block) view for structure. Report
> effect sizes and FDR, not just p-values.

## Inputs (from notebooks `21` and `22`)

- `output/21-cod-integration-crosswalk/paired-sample-metadata.csv`
- `output/22-cod-integration-feature-matrices/expression-vst-paired.csv`
- `output/22-cod-integration-feature-matrices/methylation-genebody-paired-rnaseqkey.csv`
- `output/22-cod-integration-feature-matrices/methylation-promoter-paired-rnaseqkey.csv`
- `output/22-cod-integration-feature-matrices/common-genes.txt`

## Outputs (written to `output/23-cod-integration-analysis/`)

- `gene-expr-meth-correlation.csv` — per-gene r/rho, p, BH FDR (gene-body & promoter).
- `joint-pca-scores.csv`, `joint-pca-variance.csv`
- `diablo-loadings.csv` (optional, if mixOmics is run)
- `integration-session-info.txt`

# Load packages

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)

quietly_require <- function(pkg) {
  if (suppressWarnings(requireNamespace(pkg, quietly = TRUE))) {
    library(pkg, character.only = TRUE); return(TRUE)
  }
  FALSE
}
```

# Paths

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)

in21 <- file.path(project_root, "output/21-cod-integration-crosswalk")
in22 <- file.path(project_root, "output/22-cod-integration-feature-matrices")
output_dir <- file.path(project_root, "output/23-cod-integration-analysis")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cor_method <- "spearman"   # "spearman" (rank, robust for small n) or "pearson"
fdr_alpha  <- 0.05
```

# Load matched data

```{r load-data}
paired   <- readr::read_csv(file.path(in21, "paired-sample-metadata.csv"),
                            show_col_types = FALSE)
expr     <- readr::read_csv(file.path(in22, "expression-vst-paired.csv"),
                            show_col_types = FALSE)
meth_gb  <- readr::read_csv(file.path(in22, "methylation-genebody-paired-rnaseqkey.csv"),
                            show_col_types = FALSE)
meth_pr  <- readr::read_csv(file.path(in22, "methylation-promoter-paired-rnaseqkey.csv"),
                            show_col_types = FALSE)

# Sample columns common to both assays (rnaseq_sample keys).
sample_ids <- intersect(
  setdiff(names(expr), "gene_id"),
  setdiff(names(meth_gb), "gene_id")
)
# Order samples consistently using the paired metadata.
sample_ids <- paired$rnaseq_sample[paired$rnaseq_sample %in% sample_ids]
message(glue("Samples used for integration: {length(sample_ids)}"))
stopifnot(length(sample_ids) >= 3)

# Long -> matrices (genes x samples) restricted to shared samples.
to_matrix <- function(df, cols) {
  m <- df %>% select(gene_id, all_of(cols)) %>% column_to_rownames("gene_id") %>% as.matrix()
  m
}
expr_m    <- to_matrix(expr, sample_ids)
meth_gb_m <- to_matrix(meth_gb, sample_ids)
meth_pr_m <- to_matrix(meth_pr, sample_ids)
```

# Per-gene expression–methylation correlation

```{r per-gene-correlation}
# Correlate a single gene's expression vs methylation vector across samples.
cor_one <- function(x, y, method = cor_method) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 4) return(c(estimate = NA_real_, p.value = NA_real_, n = sum(ok)))
  ct <- suppressWarnings(cor.test(x[ok], y[ok], method = method))
  c(estimate = unname(ct$estimate), p.value = ct$p.value, n = sum(ok))
}

correlate_block <- function(meth_mat, label) {
  genes <- intersect(rownames(expr_m), rownames(meth_mat))
  message(glue("[{label}] genes correlated: {length(genes)}"))
  res <- map_dfr(genes, function(g) {
    stats <- cor_one(expr_m[g, sample_ids], meth_mat[g, sample_ids])
    tibble(gene_id = g, feature = label,
           estimate = stats["estimate"], p_value = stats["p.value"], n = stats["n"])
  })
  res %>%
    mutate(fdr = p.adjust(p_value, method = "BH")) %>%
    arrange(fdr, p_value)
}

cor_genebody <- correlate_block(meth_gb_m, "gene_body")
cor_promoter <- correlate_block(meth_pr_m, "promoter")

cor_all <- bind_rows(cor_genebody, cor_promoter)
readr::write_csv(cor_all, file.path(output_dir, "gene-expr-meth-correlation.csv"))

# Quick look at the strongest relationships.
cor_all %>% filter(fdr < fdr_alpha) %>% arrange(fdr) %>% head(20)
message(glue("Genes with FDR < {fdr_alpha} (gene body): ",
             "{sum(cor_genebody$fdr < fdr_alpha, na.rm = TRUE)}"))
```

# Joint PCA on concatenated blocks

We z-score each feature within a block, concatenate the two blocks, and run a single PCA.
Restricting to the most variable features keeps the (small-n) PCA stable.

```{r joint-pca}
top_var <- function(mat, n = 1000) {
  v <- apply(mat, 1, var, na.rm = TRUE)
  mat[order(v, decreasing = TRUE)[seq_len(min(n, sum(is.finite(v))))], , drop = FALSE]
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))  # center/scale each feature (row) across samples
}

expr_top <- zscore_rows(top_var(expr_m, 1000))
meth_top <- zscore_rows(top_var(meth_gb_m, 1000))

# Drop features that became non-finite after scaling (zero variance).
expr_top <- expr_top[is.finite(rowSums(expr_top)), , drop = FALSE]
meth_top <- meth_top[is.finite(rowSums(meth_top)), , drop = FALSE]

# Prefix feature names by block, then concatenate (features x samples) -> samples x features.
rownames(expr_top) <- paste0("expr__", rownames(expr_top))
rownames(meth_top) <- paste0("meth__", rownames(meth_top))
joint <- t(rbind(expr_top, meth_top))   # samples x features

pc <- prcomp(joint, center = FALSE, scale. = FALSE)  # already z-scored
var_explained <- (pc$sdev^2) / sum(pc$sdev^2)

joint_scores <- as_tibble(pc$x[, seq_len(min(5, ncol(pc$x)))], rownames = "rnaseq_sample") %>%
  left_join(paired, by = "rnaseq_sample")
readr::write_csv(joint_scores, file.path(output_dir, "joint-pca-scores.csv"))

joint_var <- tibble(
  PC = paste0("PC", seq_along(var_explained)),
  var_explained = var_explained
) %>% slice_head(n = 10)
readr::write_csv(joint_var, file.path(output_dir, "joint-pca-variance.csv"))

joint_scores
joint_var
```

# Optional: mixOmics DIABLO multi-omics integration

DIABLO (Data Integration Analysis for Biomarker discovery using Latent components) builds
a sparse, supervised model that selects correlated features across blocks discriminating a
factor — here `temp_treatment`. With n≈10 keep components and `keepX` small, and validate
with leave-one-out.

```{r diablo}
has_mixomics <- quietly_require("mixOmics")
if (!has_mixomics) {
  message("mixOmics not installed; skipping DIABLO. ",
          "Install with BiocManager::install('mixOmics').")
} else {
  # Blocks: samples x features. Use the variable, z-scored features from above.
  X <- list(
    expression  = t(expr_top),
    methylation = t(meth_top)
  )
  Y <- paired %>%
    filter(rnaseq_sample %in% rownames(X$expression)) %>%
    arrange(match(rnaseq_sample, rownames(X$expression))) %>%
    pull(temp_treatment) %>%
    factor()

  # Full design: encourage correlation between blocks.
  design <- matrix(0.5, nrow = length(X), ncol = length(X),
                   dimnames = list(names(X), names(X)))
  diag(design) <- 0

  ncomp <- 2
  keepX <- list(expression = c(15, 15), methylation = c(15, 15))

  diablo <- mixOmics::block.splsda(
    X = X, Y = Y, ncomp = ncomp, keepX = keepX, design = design
  )

  # Leave-one-out performance (appropriate for tiny n).
  perf_diablo <- mixOmics::perf(diablo, validation = "loo")

  # Selected features / loadings per block and component.
  loadings_tbl <- map_dfr(names(X), function(blk) {
    map_dfr(seq_len(ncomp), function(cc) {
      sel <- mixOmics::selectVar(diablo, block = blk, comp = cc)
      val <- sel[[blk]]$value
      tibble(block = blk, comp = cc,
             feature = rownames(val), loading = val[, 1])
    })
  })
  readr::write_csv(loadings_tbl, file.path(output_dir, "diablo-loadings.csv"))
  loadings_tbl %>% group_by(block, comp) %>% slice_max(abs(loading), n = 5)
}
```

# Save session info

```{r session-info, eval=TRUE}
writeLines(capture.output(sessionInfo()),
           file.path(Sys.getenv("PROJECT_COD_TEMP_ROOT",
             unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)),
             "output/23-cod-integration-analysis/integration-session-info.txt"))
sessionInfo()
```

# Notes, assumptions, and TODOs

- **Direction of effect.** A negative expression–methylation correlation (esp. at
  promoters) is the classic silencing pattern; gene-body methylation often correlates
  positively with expression. Interpret sign by feature type.
- **Confounding by treatment.** With samples spanning temperatures, a gene's
  expression–methylation correlation may be driven by both responding to temperature
  rather than by a direct regulatory link. Consider partial correlation controlling for
  `temp_treatment`, or correlate within-treatment if n grows.
- **Power.** n≈10 is small. The DIABLO block is exploratory; report LOO error and avoid
  over-interpreting selected features without external validation.
- **Tissue caveat** from `21`/`22` still applies.
- **TODO (lab):** decide the primary methylation feature (gene body vs promoter) for the
  headline figures in `24`, and whether to integrate DEGs/DMR-genes only vs all common
  genes.
