---
title: "25-blood-sample-sheet"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,        # Off-repo data: run manually after checking paths
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

First notebook in the **blood RNA-seq + DNA methylation integration series** (`25`–`30`),
which implements the plan in [`21-RNAseq-methylation.md`](21-RNAseq-methylation.md).

This series is a **new, self-contained track**: blood from juvenile Pacific cod held at
two temperatures, with **the same individuals assayed by both WGBS and RNA-seq**, aligned
to a single reference (`GCF_031168955.1`). It is *separate* from the earlier liver/gill/
spleen RNA-seq work (`05`–`13`) and the earlier methylation landscape work (`18`, `20`).

**Design**

| Group | Samples (GeneticSamplingCount) | Temperature | Tissue |
|-------|--------------------------------|-------------|--------|
| Warm  | 1–36                           | 16 °C       | Blood  |
| Cold  | 44–74                          | 0 °C        | Blood  |

**The integration key is the sample ID** `{NN}B` (e.g. `01B`), which is **identical
across both assays** (verified against the Owl directory listings), so unlike the earlier
liver track there is no ambiguous crosswalk — the WGBS file `13B_*.fastq.gz` and the
RNA-seq file `13B_*.fastq.gz` are the same fish (GeneticSamplingCount 13).

**Raw data (Owl)**

- WGBS: <http://owl.fish.washington.edu/nightingales/G_macrocephalus/AN00025267/>
- RNA-seq: <http://owl.fish.washington.edu/nightingales/G_macrocephalus/AN00025268/>

**Reference genome**

- NCBI RefSeq `GCF_031168955.1` (assembly `ASM3116895v1`).
  <https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_031168955.1/>

## Outputs (written to `output/25-blood-sample-sheet/`)

- `blood-sample-metadata.csv` — one row per individual: `meth_sample` (`{NN}B`),
  `genetic_sampling_count`, `temp_treatment`, group, plus dissection covariates.
- `samplesheet-rnaseq.csv` — nf-core/Flexbar-style sheet (sample, fastq_1, fastq_2).
- `samplesheet-wgbs.csv` — nf-core/methylseq sheet (sample, fastq_1, fastq_2).
- `paired-samples.csv` — individuals with BOTH assays present on Owl (the integration set).

# Load packages

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)
library(rvest)   # parse the Owl directory index; install if needed
```

# Paths and configuration

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)

owl_base   <- "http://owl.fish.washington.edu/nightingales/G_macrocephalus"
wgbs_url   <- glue("{owl_base}/AN00025267/")
rnaseq_url <- glue("{owl_base}/AN00025268/")

# Where the FASTQs will live when staged on klone for the pipelines.
wgbs_fastq_dir   <- "/gscratch/coenv/sr320/cod-blood-bs"
rnaseq_fastq_dir <- "/gscratch/coenv/sr320/cod-blood-rna"

temp_exp_path <- file.path(project_root, "data/temp-experiment.csv")
output_dir    <- file.path(project_root, "output/25-blood-sample-sheet")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Group definition (from 21-RNAseq-methylation.md).
warm_range <- 1:36    # 16 C
cold_range <- 44:74   # 0 C
```

# Discover samples from the Owl directory listings

```{r list-owl, eval=TRUE}
# Parse an Apache directory index and return the *_1.fastq.gz basenames -> sample IDs.
list_samples <- function(url) {
  links <- rvest::read_html(url) %>%
    rvest::html_elements("a") %>%
    rvest::html_attr("href")
  r1 <- links[str_detect(links, "_1\\.fastq\\.gz$")]
  tibble(
    fastq_1 = r1,
    fastq_2 = str_replace(r1, "_1\\.fastq\\.gz$", "_2.fastq.gz"),
    sample  = str_remove(r1, "_1\\.fastq\\.gz$")   # e.g. "01B"
  )
}

wgbs_files   <- list_samples(wgbs_url)   %>% mutate(assay = "wgbs")
rnaseq_files <- list_samples(rnaseq_url) %>% mutate(assay = "rnaseq")

bind_rows(wgbs_files, rnaseq_files) %>% count(assay)
```

# Build the integrated sample table

```{r build-table, eval=TRUE}
# "01B" -> 1 ; strip trailing B and leading zeros.
count_from_id <- function(x) as.integer(str_extract(x, "^\\d+"))

assign_group <- function(n) {
  case_when(
    n %in% warm_range ~ "warm",
    n %in% cold_range ~ "cold",
    TRUE              ~ NA_character_
  )
}
assign_temp <- function(n) {
  case_when(
    n %in% warm_range ~ 16L,
    n %in% cold_range ~ 0L,
    TRUE              ~ NA_integer_
  )
}

assay_avail <- bind_rows(wgbs_files, rnaseq_files) %>%
  distinct(sample, assay) %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = assay, values_from = present, values_fill = FALSE) %>%
  rename(has_wgbs = wgbs, has_rnaseq = rnaseq)

sample_meta <- assay_avail %>%
  mutate(
    genetic_sampling_count = count_from_id(sample),
    group          = assign_group(genetic_sampling_count),
    temp_treatment = assign_temp(genetic_sampling_count),
    paired         = has_wgbs & has_rnaseq
  ) %>%
  arrange(genetic_sampling_count)

sample_meta
```

## Attach dissection covariates

```{r join-covariates, eval=TRUE}
temp_exp <- readr::read_csv(temp_exp_path, show_col_types = FALSE)

temp_exp_keyed <- temp_exp %>%
  rename(genetic_sampling_count = GeneticSamplingCount) %>%
  transmute(
    genetic_sampling_count,
    microchip_id      = `Microchip ID`,
    exp_tank          = Tank,
    exp_temp          = Temperature,
    sl_mm             = SL_mm,
    whole_body_ww_g   = WholeBodyWW_g,
    dissection_date   = DissectionDate
  ) %>%
  filter(!is.na(genetic_sampling_count)) %>%
  distinct(genetic_sampling_count, .keep_all = TRUE)

sample_meta <- sample_meta %>%
  left_join(temp_exp_keyed, by = "genetic_sampling_count")

# Sanity check: the experiment table's recorded Temperature should agree with the
# group assignment from 21-RNAseq-methylation.md. Flag any disagreements.
sample_meta %>%
  filter(!is.na(exp_temp), exp_temp != temp_treatment) %>%
  select(sample, genetic_sampling_count, temp_treatment, exp_temp)

readr::write_csv(sample_meta, file.path(output_dir, "blood-sample-metadata.csv"))
```

# Write pipeline sample sheets

```{r write-samplesheets, eval=TRUE}
# methylseq / RNA-seq sheets point at the staged klone paths.
make_sheet <- function(files, fastq_dir) {
  files %>%
    transmute(
      sample,
      fastq_1 = file.path(fastq_dir, basename(fastq_1)),
      fastq_2 = file.path(fastq_dir, basename(fastq_2))
    ) %>%
    arrange(sample)
}

readr::write_csv(make_sheet(wgbs_files,   wgbs_fastq_dir),
                 file.path(output_dir, "samplesheet-wgbs.csv"))
readr::write_csv(make_sheet(rnaseq_files, rnaseq_fastq_dir),
                 file.path(output_dir, "samplesheet-rnaseq.csv"))

paired <- sample_meta %>% filter(paired) %>%
  select(sample, genetic_sampling_count, group, temp_treatment,
         microchip_id, exp_tank)
readr::write_csv(paired, file.path(output_dir, "paired-samples.csv"))

glue("WGBS samples: {nrow(wgbs_files)} | RNA-seq samples: {nrow(rnaseq_files)} | ",
     "paired (both assays): {nrow(paired)}")
```

# Download helper (run on klone)

```{bash download, eval=FALSE}
# Stage raw reads onto klone scratch before running the pipelines.
mkdir -p /gscratch/coenv/sr320/cod-blood-bs /gscratch/coenv/sr320/cod-blood-rna

wget -r -np -nd -A "*.fastq.gz" \
  -P /gscratch/coenv/sr320/cod-blood-bs \
  http://owl.fish.washington.edu/nightingales/G_macrocephalus/AN00025267/

wget -r -np -nd -A "*.fastq.gz","*.md5" \
  -P /gscratch/coenv/sr320/cod-blood-rna \
  http://owl.fish.washington.edu/nightingales/G_macrocephalus/AN00025268/

# Verify RNA-seq checksums (Azenta provides per-sample .md5)
cd /gscratch/coenv/sr320/cod-blood-rna && md5sum -c *.md5
```

# Notes, assumptions, and TODOs

- **Shared ID = same fish.** Both Owl directories use `{NN}B` IDs; this notebook assumes
  `13B` in WGBS and `13B` in RNA-seq are the same individual (GeneticSamplingCount 13).
  This is the documented convention but should be confirmed against dissection/microchip
  records, especially for any IDs present in only one assay.
- **Group boundaries.** `warm_range`/`cold_range` encode the 1–36 = 16 °C and 44–74 = 0 °C
  split from `21-RNAseq-methylation.md`. The `exp_temp` cross-check above will surface any
  IDs whose recorded experiment temperature disagrees.
- **Tissue.** All libraries here are **blood** (note the `B` suffix), unlike the liver
  RNA-seq in `07.*`. Do not mix these count matrices.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
