---
title: "26-blood-RNAseq-align-DESeq2"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,        # Heavy compute (klone): run manually
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Second notebook in the blood integration series (`25`–`30`). It takes the **blood RNA-seq**
reads (`AN00025268`) from raw FASTQ to a **gene-level count matrix** and a **DESeq2
differential expression** result for **16 °C vs 0 °C**, aligned to `GCF_031168955.1`.

This mirrors the established liver workflow (`05` trim, `06.2` HISAT2 + featureCounts,
`07.2.2` DESeq2) but is rebuilt for blood and the two-group design so its outputs share
the `{NN}B` sample key with the methylation track (`27`).

## Inputs

- `output/25-blood-sample-sheet/samplesheet-rnaseq.csv`, `blood-sample-metadata.csv`.
- Reference: `GCF_031168955.1` genomic FASTA + GFF (staged on klone, off-repo).

## Outputs (written to `output/26-blood-RNAseq-align-DESeq2/`)

- `gene_counts_matrix.txt` — featureCounts gene-level counts (genes × `{NN}B`).
- `deseq2-results-16v0.csv` — full DESeq2 results table.
- `deseq2-DEGs-sig.csv` — significant DEGs (padj < 0.05).
- `deseq2-normalized-counts.csv` — normalized counts for downstream integration.
- `volcano-16v0.png`, `heatmap-DEGs.png`, `PCA.png`.

# Reference preparation (klone)

```{bash get-reference, eval=FALSE}
# Download GCF_031168955.1 genome + annotation from NCBI datasets.
cd /gscratch/coenv/sr320/cod-ref
datasets download genome accession GCF_031168955.1 \
  --include genome,gff3,gtf
unzip -o ncbi_dataset.zip
# Resulting files (paths used below):
#   ncbi_dataset/data/GCF_031168955.1/GCF_031168955.1_ASM3116895v1_genomic.fna
#   ncbi_dataset/data/GCF_031168955.1/genomic.gtf
```

# 1. QC and trimming

```{bash fastqc-raw, eval=FALSE}
# FastQC + MultiQC on raw reads
mkdir -p output/26-blood-RNAseq-align-DESeq2/fastqc-raw
fastqc -t 16 -o output/26-blood-RNAseq-align-DESeq2/fastqc-raw \
  /gscratch/coenv/sr320/cod-blood-rna/*.fastq.gz
multiqc output/26-blood-RNAseq-align-DESeq2/fastqc-raw \
  -o output/26-blood-RNAseq-align-DESeq2/fastqc-raw
```

```{bash trim, eval=FALSE}
# Trim with Flexbar (matching 05-cod-RNAseq-trimming conventions).
RAW=/gscratch/coenv/sr320/cod-blood-rna
TRIM=/gscratch/coenv/sr320/cod-blood-rna/trimmed
mkdir -p $TRIM

for r1 in $RAW/*_1.fastq.gz; do
  s=$(basename $r1 _1.fastq.gz)      # e.g. 13B
  flexbar \
    -r $RAW/${s}_1.fastq.gz \
    -p $RAW/${s}_2.fastq.gz \
    -a /gscratch/coenv/sr320/cod-ref/illumina_adapters.fa \
    -ap ON -qtrim TAIL -qtrim-format i1.8 -qtrim-threshold 25 \
    -m 36 -n 16 \
    -t $TRIM/${s}.flexbar_trim
done
```

# 2. Align to genome (HISAT2) and count (featureCounts)

```{bash hisat2-build, eval=FALSE}
REF=/gscratch/coenv/sr320/cod-ref/ncbi_dataset/data/GCF_031168955.1
hisat2-build -p 16 \
  $REF/GCF_031168955.1_ASM3116895v1_genomic.fna \
  /gscratch/coenv/sr320/cod-ref/GCF_031168955.1_hisat2
```

```{bash hisat2-align, eval=FALSE}
TRIM=/gscratch/coenv/sr320/cod-blood-rna/trimmed
IDX=/gscratch/coenv/sr320/cod-ref/GCF_031168955.1_hisat2
BAM=output/26-blood-RNAseq-align-DESeq2/bam
mkdir -p $BAM

for r1 in $TRIM/*.flexbar_trim_1.fastq.gz; do
  s=$(basename $r1 .flexbar_trim_1.fastq.gz)   # e.g. 13B
  hisat2 -p 16 --dta -x $IDX \
    -1 $TRIM/${s}.flexbar_trim_1.fastq.gz \
    -2 $TRIM/${s}.flexbar_trim_2.fastq.gz \
    2> $BAM/${s}.hisat2.log \
  | samtools sort -@ 16 -o $BAM/${s}.sorted.bam
  samtools index $BAM/${s}.sorted.bam
done
multiqc $BAM -o $BAM
```

```{bash featurecounts, eval=FALSE}
REF=/gscratch/coenv/sr320/cod-ref/ncbi_dataset/data/GCF_031168955.1
BAM=output/26-blood-RNAseq-align-DESeq2/bam
OUT=output/26-blood-RNAseq-align-DESeq2

# NCBI GTF: gene-level counts keyed by gene_id (e.g. LOC132...). Name BAMs as {NN}B.sorted.bam
featureCounts -T 16 -p --countReadPairs -t exon -g gene_id \
  -a $REF/genomic.gtf \
  -o $OUT/gene_counts_matrix.txt \
  $BAM/*.sorted.bam
```

# 3. Differential expression (DESeq2)

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)
quietly_require <- function(pkg) {
  if (suppressWarnings(requireNamespace(pkg, quietly = TRUE))) {
    library(pkg, character.only = TRUE); return(TRUE)
  }
  FALSE
}
```

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)
counts_path <- file.path(project_root,
  "output/26-blood-RNAseq-align-DESeq2/gene_counts_matrix.txt")
meta_path   <- file.path(project_root, "output/25-blood-sample-sheet/blood-sample-metadata.csv")
output_dir  <- file.path(project_root, "output/26-blood-RNAseq-align-DESeq2")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

padj_alpha <- 0.05
lfc_thresh <- 1     # |log2FC| > 1 for "significant" call
```

```{r import-counts}
# featureCounts: line 1 is the command comment; line 2 is the header.
fc <- readr::read_tsv(counts_path, skip = 1, show_col_types = FALSE)
annot_cols <- c("Geneid", "Chr", "Start", "End", "Strand", "Length")

count_mat <- fc %>%
  select(Geneid, all_of(setdiff(names(fc), annot_cols))) %>%
  column_to_rownames("Geneid") %>%
  as.matrix()

# Clean BAM-path column names to {NN}B sample IDs.
colnames(count_mat) <- colnames(count_mat) %>%
  basename() %>% str_remove("\\.sorted\\.bam$")
dim(count_mat)
```

```{r deseq2}
has_deseq2 <- quietly_require("DESeq2"); stopifnot(has_deseq2)

meta <- readr::read_csv(meta_path, show_col_types = FALSE) %>%
  filter(sample %in% colnames(count_mat)) %>%
  mutate(temp_treatment = factor(temp_treatment, levels = c(0, 16))) %>%  # 0 = reference
  column_to_rownames("sample")
count_mat <- count_mat[, rownames(meta), drop = FALSE]

# Pre-filter low-count genes.
keep <- rowSums(count_mat >= 10) >= 3
count_filt <- count_mat[keep, , drop = FALSE]
message(glue("Genes kept: {nrow(count_filt)} / {nrow(count_mat)}"))

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = count_filt,
  colData   = meta,
  design    = ~ temp_treatment
)
dds <- DESeq2::DESeq(dds)

# Contrast: 16 C vs 0 C (positive log2FC = higher in warm).
res <- DESeq2::results(dds, contrast = c("temp_treatment", "16", "0"))
res_df <- as_tibble(as.data.frame(res), rownames = "gene_id") %>% arrange(padj)
readr::write_csv(res_df, file.path(output_dir, "deseq2-results-16v0.csv"))

degs <- res_df %>% filter(!is.na(padj), padj < padj_alpha, abs(log2FoldChange) > lfc_thresh)
readr::write_csv(degs, file.path(output_dir, "deseq2-DEGs-sig.csv"))
message(glue("Significant DEGs (padj<{padj_alpha}, |log2FC|>{lfc_thresh}): {nrow(degs)}"))

# Normalized counts + VST for downstream integration.
norm_counts <- DESeq2::counts(dds, normalized = TRUE)
readr::write_csv(as_tibble(norm_counts, rownames = "gene_id"),
                 file.path(output_dir, "deseq2-normalized-counts.csv"))
vsd <- DESeq2::vst(dds, blind = TRUE)
readr::write_csv(as_tibble(SummarizedExperiment::assay(vsd), rownames = "gene_id"),
                 file.path(output_dir, "deseq2-vst.csv"))
```

# 4. Visualizations

```{r figures}
library(ggplot2)

# Volcano
volc <- res_df %>% mutate(
  sig = !is.na(padj) & padj < padj_alpha & abs(log2FoldChange) > lfc_thresh)
p_volc <- ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = sig)) +
  geom_point(alpha = 0.5, size = 1.6) +
  scale_color_manual(values = c(`TRUE` = "#B2182B", `FALSE` = "grey70")) +
  geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dotted") +
  labs(title = "Blood RNA-seq: 16 °C vs 0 °C", x = "log2 fold change (16 vs 0)",
       y = expression(-log[10](p))) +
  theme_bw()
ggsave(file.path(output_dir, "volcano-16v0.png"), p_volc, width = 7, height = 5, dpi = 300)

# PCA
pca <- DESeq2::plotPCA(vsd, intgroup = "temp_treatment", returnData = TRUE)
pv <- round(100 * attr(pca, "percentVar"))
p_pca <- ggplot(pca, aes(PC1, PC2, color = temp_treatment, label = name)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(`0` = "#2166AC", `16` = "#B2182B"), name = "Temp (°C)") +
  labs(x = glue("PC1 ({pv[1]}%)"), y = glue("PC2 ({pv[2]}%)"),
       title = "Blood RNA-seq PCA") + theme_bw()
ggsave(file.path(output_dir, "PCA.png"), p_pca, width = 7, height = 5, dpi = 300)

# Heatmap of significant DEGs
if (quietly_require("pheatmap") && nrow(degs) > 1) {
  m <- SummarizedExperiment::assay(vsd)[degs$gene_id, , drop = FALSE]
  m <- t(scale(t(m)))
  ann <- as.data.frame(meta["temp_treatment"])
  pheatmap::pheatmap(m, annotation_col = ann, show_rownames = FALSE,
    main = "Significant DEGs (VST, row z-score)",
    filename = file.path(output_dir, "heatmap-DEGs.png"), width = 8, height = 8)
}
```

# Notes, assumptions, and TODOs

- **Reference consistency.** Use the *same* `GCF_031168955.1` assembly and chromosome
  accessions (`NC_0823..`) as the methylation track (`27`) so gene IDs and coordinates
  join cleanly in `28`.
- **gene_id attribute.** NCBI GTF uses `gene_id` like `LOC132...`; confirm the
  `featureCounts -g gene_id` column matches the `gene_id` used when summarizing
  methylation to genes in `28`.
- **Alternative DE.** An edgeR path (cf. `13.0.0`) can be added if preferred; DESeq2 is the
  headline here for consistency with `07.2.2`.
- **Batch/covariates.** If tank or extraction batch is a concern, add it to the design
  (`~ tank + temp_treatment`) — covariates are available in `blood-sample-metadata.csv`.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
