---
title: "27-blood-WGBS-methylation-DMR"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,        # Heavy compute (klone): run manually
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Third notebook in the blood integration series (`25`–`30`). It takes the **blood WGBS**
reads (`AN00025267`) through the nf-core/methylseq (Bismark) pipeline to per-CpG coverage,
then calls **differentially methylated CpGs (DMCs) and regions (DMRs)** between
**16 °C and 0 °C** and summarizes methylation to **gene-level features** for integration.

The methylseq pipeline is already configured for this genome in the repo (`16.2-genome-prep`,
`17.nextflow`, `17.config`, `17.samplesheet.csv`). This notebook reuses that setup with the
blood sample sheet from `25`, then adds DMR calling (methylKit) and per-gene summaries.

## Inputs

- `output/25-blood-sample-sheet/samplesheet-wgbs.csv`, `blood-sample-metadata.csv`.
- Reference `GCF_031168955.1` FASTA + GFF/GTF (same as `26`).
- methylseq `coverage2cytosine` `*.cov.gz` outputs (off-repo on klone/Owl).

## Outputs (written to `output/27-blood-WGBS-methylation-DMR/`)

- `methylation-landscape-summary.csv` — global %mCpG per sample.
- `DMCs-16v0.csv` — per-CpG differential methylation (methylKit).
- `DMRs-16v0.csv` — differentially methylated regions (tiled or methylKit regions).
- `genebody-methylation.csv`, `promoter-methylation.csv` — gene × `{NN}B` matrices.
- `gene-methylation-diff.csv` — per-gene mean Δmethylation (16 − 0) + significance.

# Run methylseq (klone)

```{bash methylseq, eval=FALSE}
# Reuse the repo's nf-core/methylseq config with the BLOOD sample sheet from 25.
# (cf. code/17.nextflow). coverage2cytosine output is required for CpG-resolution.
nextflow run nf-core/methylseq \
  -c code/17.config \
  --input output/25-blood-sample-sheet/samplesheet-wgbs.csv \
  --outdir output/27-blood-WGBS-methylation-DMR/methylseq \
  --fasta /gscratch/coenv/sr320/cod-ref/GCF_031168955.1_ASM3116895v1_genomic.fna \
  --cytosine_report \
  -resume -with-trace -with-report nf_report.html -with-timeline nf_timeline.html
```

# Load packages

```{r load-packages, eval=TRUE}
library(tidyverse)
library(glue)
library(vroom)
quietly_require <- function(pkg) {
  if (suppressWarnings(requireNamespace(pkg, quietly = TRUE))) {
    library(pkg, character.only = TRUE); return(TRUE)
  }
  FALSE
}
```

# Paths and configuration

```{r paths, eval=TRUE}
project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)

# Set to wherever the coverage2cytosine *.cov.gz files live in your environment.
cov_dir  <- file.path(project_root,
  "output/27-blood-WGBS-methylation-DMR/methylseq/coverage2cytosine/coverage")
meta_path <- file.path(project_root, "output/25-blood-sample-sheet/blood-sample-metadata.csv")
gtf_path  <- file.path(project_root, "data/genomic.gtf")   # GCF_031168955.1 (off-repo)

output_dir <- file.path(project_root, "output/27-blood-WGBS-methylation-DMR")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

min_cov           <- 5L       # min per-sample CpG depth
min_per_group     <- 3L       # min samples per group covering a CpG for DMC test
qvalue_cutoff     <- 0.01     # methylKit DMC q-value
meth_diff_cutoff  <- 25       # methylKit DMC % difference
tile_window       <- 1000L    # DMR tile width (bp)
promoter_upstream <- 1000L
promoter_downstream <- 200L
min_cpgs_per_gene <- 3L
```

```{r load-meta, eval=TRUE}
meta <- readr::read_csv(meta_path, show_col_types = FALSE) %>%
  filter(has_wgbs)
# Map sample -> group treatment (0 = cold, 1 = warm) for methylKit.
meta_meth <- meta %>% transmute(sample, temp_treatment,
                                treatment = if_else(temp_treatment == 16, 1L, 0L))
meta_meth
```

# 1. Methylation landscape summary

```{r landscape}
cov_files <- list.files(cov_dir, pattern = "\\.cov(\\.gz)?$", full.names = TRUE)
stopifnot(length(cov_files) > 0)
cov_label <- function(p) sub("\\..*$", "", basename(p))   # "13B.*.cov.gz" -> "13B"

read_cov <- function(path, min_coverage = min_cov) {
  vroom::vroom(path, delim = "\t",
    col_names = c("chr", "start", "end", "perc_meth", "count_meth", "count_unmeth"),
    col_types = list(chr = "c", start = "i", end = "i",
                     perc_meth = "d", count_meth = "i", count_unmeth = "i"),
    progress = FALSE) %>%
    mutate(depth = count_meth + count_unmeth) %>%
    filter(depth >= min_coverage)
}

landscape <- map_dfr(cov_files, function(p) {
  d <- read_cov(p)
  tibble(sample = cov_label(p),
         n_cpg = nrow(d),
         global_pct_meth = 100 * sum(d$count_meth) / sum(d$count_meth + d$count_unmeth))
}) %>% left_join(meta_meth, by = "sample")
readr::write_csv(landscape, file.path(output_dir, "methylation-landscape-summary.csv"))
landscape
```

# 2. Differential methylation — CpGs and regions (methylKit)

```{r methylkit}
has_mk <- quietly_require("methylKit"); stopifnot(has_mk)

# Restrict to WGBS samples present in cov_dir, ordered with treatment vector.
cov_index <- tibble(path = cov_files, sample = map_chr(cov_files, cov_label)) %>%
  inner_join(meta_meth, by = "sample") %>%
  arrange(treatment, sample)

obj <- methylKit::methRead(
  location   = as.list(cov_index$path),
  sample.id  = as.list(cov_index$sample),
  assembly   = "GCF_031168955.1",
  treatment  = cov_index$treatment,
  context    = "CpG",
  pipeline   = "bismarkCoverage",
  mincov     = min_cov
)

# Per-CpG DMCs: require coverage in most samples of each group.
filt   <- methylKit::filterByCoverage(obj, lo.count = min_cov, hi.perc = 99.9)
united <- methylKit::unite(filt, destrand = FALSE,
                           min.per.group = as.integer(min_per_group))
diff   <- methylKit::calculateDiffMeth(united)

dmc <- methylKit::getMethylDiff(diff, difference = meth_diff_cutoff,
                                qvalue = qvalue_cutoff)
dmc_df <- methylKit::getData(dmc) %>% as_tibble()
readr::write_csv(dmc_df, file.path(output_dir, "DMCs-16v0.csv"))
message(glue("DMCs (q<{qvalue_cutoff}, |diff|>{meth_diff_cutoff}%): {nrow(dmc_df)}"))

# DMRs via tiling.
tiles  <- methylKit::tileMethylCounts(obj, win.size = tile_window,
                                      step.size = tile_window, cov.bases = min_cpgs_per_gene)
t_unite <- methylKit::unite(tiles, destrand = FALSE,
                            min.per.group = as.integer(min_per_group))
t_diff  <- methylKit::calculateDiffMeth(t_unite)
dmr     <- methylKit::getMethylDiff(t_diff, difference = meth_diff_cutoff,
                                    qvalue = qvalue_cutoff)
dmr_df  <- methylKit::getData(dmr) %>% as_tibble()
readr::write_csv(dmr_df, file.path(output_dir, "DMRs-16v0.csv"))
message(glue("DMRs ({tile_window}bp tiles): {nrow(dmr_df)}"))
```

# 3. Gene-level methylation summaries

```{r gene-features}
has_gr <- quietly_require("GenomicRanges"); has_gf <- quietly_require("GenomicFeatures")
stopifnot(has_gr, has_gf)

make_txdb <- function(gtf) {
  if (quietly_require("txdbmaker")) txdbmaker::makeTxDbFromGFF(gtf, format = "gtf")
  else GenomicFeatures::makeTxDbFromGFF(gtf, format = "gtf")
}
txdb <- make_txdb(gtf_path)
gene_bodies   <- GenomicFeatures::genes(txdb)
gene_promoters <- GenomicRanges::promoters(gene_bodies,
  upstream = promoter_upstream, downstream = promoter_downstream)

# Coverage-weighted methylation fraction per gene, per sample.
summarize_sample <- function(cov_df, feature_gr, min_cpgs = min_cpgs_per_gene) {
  cpg_gr <- GenomicRanges::GRanges(cov_df$chr,
    IRanges::IRanges(cov_df$start, cov_df$start),
    count_meth = cov_df$count_meth, count_unmeth = cov_df$count_unmeth)
  hits <- GenomicRanges::findOverlaps(cpg_gr, feature_gr, ignore.strand = TRUE)
  if (length(hits) == 0) return(tibble(gene_id = character(), meth = numeric()))
  tibble(
    gene_id      = names(feature_gr)[S4Vectors::subjectHits(hits)],
    count_meth   = cov_df$count_meth[S4Vectors::queryHits(hits)],
    count_unmeth = cov_df$count_unmeth[S4Vectors::queryHits(hits)]
  ) %>%
    group_by(gene_id) %>%
    summarise(n_cpg = n(),
              meth = sum(count_meth) / sum(count_meth + count_unmeth), .groups = "drop") %>%
    filter(n_cpg >= min_cpgs) %>% select(gene_id, meth)
}

build_matrix <- function(feature_gr) {
  cov_index %>%
    mutate(data = map(path, ~ summarize_sample(read_cov(.x), feature_gr))) %>%
    select(sample, data) %>% unnest(data) %>%
    pivot_wider(names_from = sample, values_from = meth)
}

meth_genebody <- build_matrix(gene_bodies)
meth_promoter <- build_matrix(gene_promoters)
readr::write_csv(meth_genebody, file.path(output_dir, "genebody-methylation.csv"))
readr::write_csv(meth_promoter, file.path(output_dir, "promoter-methylation.csv"))
dim(meth_genebody); dim(meth_promoter)
```

```{r gene-diff}
# Per-gene mean methylation by group + Δ(16 − 0) and a simple t-test (gene-body).
warm_ids <- meta_meth$sample[meta_meth$treatment == 1]
cold_ids <- meta_meth$sample[meta_meth$treatment == 0]

gene_diff <- meth_genebody %>%
  rowwise() %>%
  mutate(
    mean_warm = mean(c_across(any_of(warm_ids)), na.rm = TRUE),
    mean_cold = mean(c_across(any_of(cold_ids)), na.rm = TRUE),
    delta     = mean_warm - mean_cold,
    p_value   = tryCatch(
      t.test(c_across(any_of(warm_ids)), c_across(any_of(cold_ids)))$p.value,
      error = function(e) NA_real_)
  ) %>%
  ungroup() %>%
  select(gene_id, mean_warm, mean_cold, delta, p_value) %>%
  mutate(fdr = p.adjust(p_value, "BH")) %>%
  arrange(fdr)
readr::write_csv(gene_diff, file.path(output_dir, "gene-methylation-diff.csv"))
message(glue("Differentially methylated genes (FDR<0.05): ",
             "{sum(gene_diff$fdr < 0.05, na.rm = TRUE)}"))
```

# Notes, assumptions, and TODOs

- **Seqlevels must match.** GTF and `.cov` chromosome names are both RefSeq `NC_0823..`
  accessions from `GCF_031168955.1`; if they ever diverge, harmonize with
  `GenomeInfoDb::seqlevelsStyle()` before `findOverlaps`.
- **DMR method.** Tiling (1 kb) via methylKit is the default here; `dmrseq` or `DSS` are
  reasonable alternatives if you want smoothing / region detection rather than fixed tiles.
- **Power.** Group sizes are modest; report effect size (% difference) alongside q-values
  and avoid over-interpreting individual DMCs.
- **Feature definition.** Gene-body and promoter (±1 kb/+200 bp) are defaults; CpG-island
  or first-exon summaries can reuse `summarize_sample()` with a different `GRanges`.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
