---
title: "28-blood-integration-feature-matrices"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Fourth notebook in the blood integration series (`25`–`30`). It assembles **matched
feature matrices** so the integration in `29` operates on the same genes and the same
**paired individuals** across both assays.

Because blood WGBS and RNA-seq share the `{NN}B` sample ID (verified in `25`), no
crosswalk is needed — matrices are joined directly on sample ID and on `gene_id`.

Three matrices, restricted to paired samples:

1. **Gene expression** — DESeq2 VST from `26` (genes × `{NN}B`).
2. **Gene-body methylation** — weighted CpG fraction from `27`.
3. **Promoter methylation** — same, over TSS-flanking windows.

## Inputs

- `output/25-blood-sample-sheet/paired-samples.csv`
- `output/26-blood-RNAseq-align-DESeq2/deseq2-vst.csv`
- `output/27-blood-WGBS-methylation-DMR/genebody-methylation.csv`
- `output/27-blood-WGBS-methylation-DMR/promoter-methylation.csv`

## Outputs (written to `output/28-blood-integration-feature-matrices/`)

- `expression-vst-paired.csv`, `methylation-genebody-paired.csv`,
  `methylation-promoter-paired.csv`
- `common-genes.txt`, `feature-matrix-summary.csv`

# Load packages and paths

```{r load, eval=TRUE}
library(tidyverse)
library(glue)

project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)
in25 <- file.path(project_root, "output/25-blood-sample-sheet")
in26 <- file.path(project_root, "output/26-blood-RNAseq-align-DESeq2")
in27 <- file.path(project_root, "output/27-blood-WGBS-methylation-DMR")
output_dir <- file.path(project_root, "output/28-blood-integration-feature-matrices")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

```{r read-inputs}
paired  <- readr::read_csv(file.path(in25, "paired-samples.csv"), show_col_types = FALSE)
expr    <- readr::read_csv(file.path(in26, "deseq2-vst.csv"), show_col_types = FALSE)
meth_gb <- readr::read_csv(file.path(in27, "genebody-methylation.csv"), show_col_types = FALSE)
meth_pr <- readr::read_csv(file.path(in27, "promoter-methylation.csv"), show_col_types = FALSE)

# Paired samples present in BOTH matrices (shared {NN}B key).
paired_ids <- paired$sample
expr_ids   <- intersect(paired_ids, setdiff(names(expr), "gene_id"))
meth_ids   <- intersect(paired_ids, setdiff(names(meth_gb), "gene_id"))
shared_ids <- intersect(expr_ids, meth_ids)
message(glue("Paired samples usable in both assays: {length(shared_ids)}"))
stopifnot(length(shared_ids) >= 3)
```

# Restrict matrices to shared samples and common genes

```{r restrict}
sel <- function(df) df %>% select(gene_id, all_of(shared_ids))
expr_p    <- sel(expr)
meth_gb_p <- sel(meth_gb)
meth_pr_p <- sel(meth_pr)

readr::write_csv(expr_p,    file.path(output_dir, "expression-vst-paired.csv"))
readr::write_csv(meth_gb_p, file.path(output_dir, "methylation-genebody-paired.csv"))
readr::write_csv(meth_pr_p, file.path(output_dir, "methylation-promoter-paired.csv"))

common_genes <- intersect(expr_p$gene_id, meth_gb_p$gene_id)
writeLines(common_genes, file.path(output_dir, "common-genes.txt"))

feature_summary <- tibble(
  matrix = c("expression_vst", "methylation_genebody", "methylation_promoter", "common_genes"),
  n_features = c(nrow(expr_p), nrow(meth_gb_p), nrow(meth_pr_p), length(common_genes)),
  n_samples  = c(length(shared_ids), length(shared_ids), length(shared_ids), NA_integer_)
)
readr::write_csv(feature_summary, file.path(output_dir, "feature-matrix-summary.csv"))
feature_summary
```

# Notes

- **Single shared key.** No `{NN}B ↔ sample_{N}` translation is required for blood; both
  assays use `{NN}B`. If an individual is missing from one matrix it is simply dropped
  from the integration set.
- **No feature pre-selection here.** Full normalized matrices are written; DEG/DMR-based
  selection happens in `29`.
- **gene_id consistency.** Expression `gene_id` (featureCounts `-g gene_id`, `26`) must be
  the same identifier used when summarizing methylation to genes (`27`); both derive from
  the `GCF_031168955.1` GTF.

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
