---
title: "29-blood-integration-analysis"
author: "Roberts Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
always_allow_html: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = ""
)
```

# Overview

Fifth notebook in the blood integration series (`25`–`30`). It performs the core
**expression ↔ methylation integration** to address the project question: *how does DNA
methylation relate to gene expression, and which thermally-responsive genes appear to be
under epigenetic control?*

Four analyses:

1. **Genome-wide relationship.** Across all genes, relate mean methylation (promoter and
   gene-body) to mean expression level — establishes the baseline epigenetic pattern
   (expected: promoter meth ↓ expression; gene-body meth ↑ expression).
2. **Per-gene correlation** of expression vs methylation across paired individuals
   (BH FDR), separately for gene-body and promoter.
3. **DEG × DMG overlap (the headline).** Cross the temperature DEGs (`26`) with the
   differentially methylated genes (`27`) into a four-quadrant table (e.g. promoter-
   hypermethylated + down-regulated) — the candidate epigenetically-controlled thermal
   response genes.
4. **Joint multivariate view** (PCA; optional mixOmics DIABLO) to see whether the combined
   omics separate 16 °C vs 0 °C.

## Inputs

- `output/25-blood-sample-sheet/paired-samples.csv`
- `output/26-blood-RNAseq-align-DESeq2/deseq2-results-16v0.csv`
- `output/27-blood-WGBS-methylation-DMR/gene-methylation-diff.csv`
- `output/28-blood-integration-feature-matrices/*` (paired matrices, common-genes)

## Outputs (written to `output/29-blood-integration-analysis/`)

- `genome-wide-meth-expr.csv` — binned methylation vs mean expression.
- `gene-expr-meth-correlation.csv` — per-gene r/rho, p, FDR (gene-body & promoter).
- `deg-dmg-overlap.csv` — DEG × DMG quadrant assignments for candidate genes.
- `joint-pca-scores.csv`, `joint-pca-variance.csv`, `diablo-loadings.csv` (optional).

# Load packages and paths

```{r load, eval=TRUE}
library(tidyverse)
library(glue)
quietly_require <- function(pkg) {
  if (suppressWarnings(requireNamespace(pkg, quietly = TRUE))) {
    library(pkg, character.only = TRUE); return(TRUE)
  }
  FALSE
}

project_root <- Sys.getenv(
  "PROJECT_COD_TEMP_ROOT",
  unset = normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
)
in25 <- file.path(project_root, "output/25-blood-sample-sheet")
in26 <- file.path(project_root, "output/26-blood-RNAseq-align-DESeq2")
in27 <- file.path(project_root, "output/27-blood-WGBS-methylation-DMR")
in28 <- file.path(project_root, "output/28-blood-integration-feature-matrices")
output_dir <- file.path(project_root, "output/29-blood-integration-analysis")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cor_method <- "spearman"
fdr_alpha  <- 0.05
padj_alpha <- 0.05
lfc_thresh <- 1
```

```{r read-inputs}
paired   <- readr::read_csv(file.path(in25, "paired-samples.csv"), show_col_types = FALSE)
deg      <- readr::read_csv(file.path(in26, "deseq2-results-16v0.csv"), show_col_types = FALSE)
gene_meth_diff <- readr::read_csv(file.path(in27, "gene-methylation-diff.csv"),
                                  show_col_types = FALSE)
expr     <- readr::read_csv(file.path(in28, "expression-vst-paired.csv"), show_col_types = FALSE)
meth_gb  <- readr::read_csv(file.path(in28, "methylation-genebody-paired.csv"),
                            show_col_types = FALSE)
meth_pr  <- readr::read_csv(file.path(in28, "methylation-promoter-paired.csv"),
                            show_col_types = FALSE)

sample_ids <- intersect(setdiff(names(expr), "gene_id"), setdiff(names(meth_gb), "gene_id"))
sample_ids <- paired$sample[paired$sample %in% sample_ids]
to_matrix <- function(df) df %>% select(gene_id, all_of(sample_ids)) %>%
  column_to_rownames("gene_id") %>% as.matrix()
expr_m    <- to_matrix(expr)
meth_gb_m <- to_matrix(meth_gb)
meth_pr_m <- to_matrix(meth_pr)
```

# 1. Genome-wide methylation–expression relationship

```{r genome-wide}
mean_expr <- tibble(gene_id = rownames(expr_m), mean_expr = rowMeans(expr_m, na.rm = TRUE))
mean_gb   <- tibble(gene_id = rownames(meth_gb_m), gene_body = rowMeans(meth_gb_m, na.rm = TRUE))
mean_pr   <- tibble(gene_id = rownames(meth_pr_m), promoter  = rowMeans(meth_pr_m, na.rm = TRUE))

gw <- mean_expr %>% left_join(mean_gb, by = "gene_id") %>% left_join(mean_pr, by = "gene_id")

# Bin by methylation decile and summarize expression — the canonical curve.
gw_binned <- gw %>%
  pivot_longer(c(gene_body, promoter), names_to = "feature", values_to = "methylation") %>%
  filter(is.finite(methylation), is.finite(mean_expr)) %>%
  group_by(feature) %>%
  mutate(meth_bin = ntile(methylation, 10)) %>%
  group_by(feature, meth_bin) %>%
  summarise(mean_expr = mean(mean_expr), n = n(), .groups = "drop")
readr::write_csv(gw_binned, file.path(output_dir, "genome-wide-meth-expr.csv"))
gw_binned
```

# 2. Per-gene expression–methylation correlation

```{r per-gene-corr}
cor_one <- function(x, y, method = cor_method) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 4) return(c(estimate = NA_real_, p.value = NA_real_, n = sum(ok)))
  ct <- suppressWarnings(cor.test(x[ok], y[ok], method = method))
  c(estimate = unname(ct$estimate), p.value = ct$p.value, n = sum(ok))
}
correlate_block <- function(meth_mat, label) {
  genes <- intersect(rownames(expr_m), rownames(meth_mat))
  map_dfr(genes, function(g) {
    s <- cor_one(expr_m[g, sample_ids], meth_mat[g, sample_ids])
    tibble(gene_id = g, feature = label,
           estimate = s["estimate"], p_value = s["p.value"], n = s["n"])
  }) %>% mutate(fdr = p.adjust(p_value, "BH")) %>% arrange(fdr, p_value)
}
cor_all <- bind_rows(correlate_block(meth_gb_m, "gene_body"),
                     correlate_block(meth_pr_m, "promoter"))
readr::write_csv(cor_all, file.path(output_dir, "gene-expr-meth-correlation.csv"))
cor_all %>% filter(fdr < fdr_alpha) %>% head(20)
```

# 3. DEG × DMG overlap (candidate epigenetically-controlled genes)

```{r deg-dmg}
deg_tagged <- deg %>%
  transmute(gene_id,
            log2FoldChange,
            deg_sig = !is.na(padj) & padj < padj_alpha & abs(log2FoldChange) > lfc_thresh,
            expr_dir = if_else(log2FoldChange > 0, "up_warm", "down_warm"))

dmg_tagged <- gene_meth_diff %>%
  transmute(gene_id,
            delta,                              # mean gene-body meth (16 - 0)
            dmg_sig = !is.na(fdr) & fdr < fdr_alpha,
            meth_dir = if_else(delta > 0, "hyper_warm", "hypo_warm"))

overlap <- deg_tagged %>%
  inner_join(dmg_tagged, by = "gene_id") %>%
  mutate(quadrant = paste(expr_dir, meth_dir, sep = " / "),
         candidate = deg_sig & dmg_sig)

readr::write_csv(overlap, file.path(output_dir, "deg-dmg-overlap.csv"))

# Quadrant counts among genes significant in BOTH assays.
overlap %>% filter(candidate) %>% count(quadrant, sort = TRUE)
message(glue("DEG ∩ DMG candidates: {sum(overlap$candidate)}"))
```

# 4. Joint PCA

```{r joint-pca}
top_var <- function(mat, n = 1000) {
  v <- apply(mat, 1, var, na.rm = TRUE)
  mat[order(v, decreasing = TRUE)[seq_len(min(n, sum(is.finite(v))))], , drop = FALSE]
}
zrows <- function(mat) t(scale(t(mat)))
expr_top <- zrows(top_var(expr_m)); meth_top <- zrows(top_var(meth_gb_m))
expr_top <- expr_top[is.finite(rowSums(expr_top)), , drop = FALSE]
meth_top <- meth_top[is.finite(rowSums(meth_top)), , drop = FALSE]
rownames(expr_top) <- paste0("expr__", rownames(expr_top))
rownames(meth_top) <- paste0("meth__", rownames(meth_top))
joint <- t(rbind(expr_top, meth_top))

pc <- prcomp(joint, center = FALSE, scale. = FALSE)
ve <- (pc$sdev^2) / sum(pc$sdev^2)
joint_scores <- as_tibble(pc$x[, seq_len(min(5, ncol(pc$x)))], rownames = "sample") %>%
  left_join(paired, by = "sample")
readr::write_csv(joint_scores, file.path(output_dir, "joint-pca-scores.csv"))
readr::write_csv(tibble(PC = paste0("PC", seq_along(ve)), var_explained = ve) %>%
                   slice_head(n = 10),
                 file.path(output_dir, "joint-pca-variance.csv"))
joint_scores
```

# 5. Optional: mixOmics DIABLO

```{r diablo}
has_mixomics <- quietly_require("mixOmics")
if (!has_mixomics) {
  message("mixOmics not installed; skipping DIABLO.")
} else {
  X <- list(expression = t(expr_top), methylation = t(meth_top))
  Y <- paired %>% filter(sample %in% rownames(X$expression)) %>%
    arrange(match(sample, rownames(X$expression))) %>% pull(group) %>% factor()
  design <- matrix(0.5, length(X), length(X), dimnames = list(names(X), names(X)))
  diag(design) <- 0
  diablo <- mixOmics::block.splsda(X = X, Y = Y, ncomp = 2,
    keepX = list(expression = c(15, 15), methylation = c(15, 15)), design = design)
  loadings_tbl <- map_dfr(names(X), function(blk) map_dfr(1:2, function(cc) {
    val <- mixOmics::selectVar(diablo, block = blk, comp = cc)[[blk]]$value
    tibble(block = blk, comp = cc, feature = rownames(val), loading = val[, 1])
  }))
  readr::write_csv(loadings_tbl, file.path(output_dir, "diablo-loadings.csv"))
  loadings_tbl %>% group_by(block, comp) %>% slice_max(abs(loading), n = 5)
}
```

```{r session-info, eval=TRUE}
writeLines(capture.output(sessionInfo()),
           file.path(output_dir, "integration-session-info.txt"))
sessionInfo()
```

# Notes, assumptions, and TODOs

- **Direction of effect.** Negative promoter-meth ↔ expression is the classic silencing
  pattern; gene-body methylation often correlates positively with expression. Interpret
  the sign by feature type, and read the DEG×DMG quadrants accordingly (the
  `down_warm / hyper_warm` and `up_warm / hypo_warm` promoter quadrants are the cleanest
  "epigenetic control" candidates).
- **Confounding by treatment.** A per-gene expression–methylation correlation may reflect
  both responding to temperature rather than a direct regulatory link. Consider partial
  correlation controlling for group, or correlating within-group as n allows.
- **Power.** Treat per-gene results as hypothesis-generating; lean on the DEG×DMG overlap
  and the multivariate view for structure.
- **Functional follow-up.** Pass the candidate gene set to GO/enrichment (reuse the
  `08.*` / DAVID approach) — this can be a short `29.1` notebook or a section in `30`.

