---
title: "Phenotypes, P. cod temp experiment"
author: "Laura Spencer"
date: "2024-07-29"
output:
  html_document: default
  pdf_document: default
---

Juvenile Pacific cod were tagged, acclimated to laboratory conditions, then gradually acclimated to four experimental temperatures (0, 5, 9, 16). After 6 weeks in treatments, fish were sacrificed, liver and fin tissues were collected for RNASeq and lcWGS, respectively, and measurements taken of fish length, wet weight, and liver weight. 

On 11/21/22 160 tagged fish entered experimental tanks for acclimation, 40 per treatment (n=10 fish / tank, 4 tanks/treatment). On 12/28/22 temperatures began to slowly increase until all target experimental temperatures (0C, 5C, 9C, 16C) were reached on 1/8/23. Of the 40 fish per treatment, all survived the 0C, 5C, and 9C treatments. Four fish died in the 16degC treatment on 1/22/23 (tank 1), 1/24/23 (tank 2), 1/29/23 (tank 4), and 2/1/23 (tank 2). All survivors were euthanized/sampled on 2/8 & 2/9. Standard length and wet weight were 1) collected before tank acclimation, 2) before experimental temperature, and 3) at treatment termination so that growth and condition could be assessed prior to and during experimental treatments. Liver wet weight and tissues for sequencing were collected at treatment termination. 

In this notebook I look at effects of temperature on distributions of growth rate (length, weight) body condition index (wet weight / length), liver condition (i.e. hepato-somatic index, liver wet weight / whole body wet weight), and lipids.

### Load libraries and source scripts 

```{r, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("tidyverse", "readxl", "janitor", "purrr", "ggpubr", "googlesheets4", "plotly", "lubridate", "scales", "knitr", "cowplot")

# Load all required libraries
all.packages <- c(list.of.packages)
lapply(all.packages, FUN = function(X) {
  do.call("require", list(X))
})

`%!in%` = Negate(`%in%`)
```

### Load data and calculate per-fish metrics 

Specific growth rates (SGR) is calculated for each fish based on wet weights and standard length using the below equation, where M1 and M2 are measurements (standard length, wet weight) taken before (t1) and after (t2) the acclimation period to assess baseline growth rates, and before and after experimental temperature treatments to assess effects of temperature (Table 1).

SGR(standard length, wet weight) = 100 * (e^g-1) where g = (ln(M2) - ln(M1)) / t2-t1 

Fulton’s condition index (Kwet) was approximated as K(wet) = 100 * wet total weight (g) / (standard length (mm)/10)^3

The hepato-somatic index (HSI) was calculated as HSI =100 x wet liver weight (g)wet total weight (g). 

```{r, message=F, warning=F}
load(file = "../rnaseq/aligned/counts.ts")

phenotypes <- read_excel("../data/Pcod Temp Growth experiment 2022-23 DATA.xlsx", sheet = "AllData") %>% clean_names() %>%
  mutate(acc_t1=as.Date("2022-11-21", format="%Y-%m-%d"), exp_t1=as.Date("2022-12-27", format= "%Y-%m-%d"), dissection_date=as.Date(dissection_date, format= "%Y-%m-%d")) %>%
  mutate_at(c("tank", "temperature", "microchip_id", "genetic_sampling_count"), factor) %>%
  dplyr::rename("sl_final"="sl_mm", "wwt_final"="whole_body_ww_g") %>% 
  mutate(SGR.sl.accl=round(100*(exp((log(sl_12272022)-log(sl_11212022))/(as.numeric(exp_t1-acc_t1)))-1), digits=5),
         SGR.sl.trt=round(100*(exp((log(sl_final)-log(sl_12272022))/(as.numeric(dissection_date-exp_t1)))-1), digits=5),
         SGR.ww.accl=round(100*(exp((log(wwt_12272022)-log(wwt_11212022))/(as.numeric(exp_t1-acc_t1)))-1), digits=5),
         SGR.ww.trt=round(100*(exp((log(wwt_final)-log(wwt_12272022))/(as.numeric(dissection_date-exp_t1)))-1), digits=5),
         Kwet=100*(wwt_final/(sl_final/10)^3),
         hsi=100*total_liver_ww_mg/(wwt_final),
         mort=as.factor(case_when(is.na(mort_date) ~ "No", TRUE ~ "Yes")),
         rna=case_when(genetic_sampling_count %in% gsub("sample_", "", rownames(counts.ts)) ~ "Yes", TRUE ~ "No"),
         measure_date=case_when(is.na(mort_date) ~ ymd(dissection_date), TRUE ~ mort_date)) %>%
  mutate(days_growth = as.numeric(difftime(ymd(measure_date), ymd("2022-12-27"), units = "days")))

phenotypes %>% head()
```

```{r, message=F, warning=F, results=F}
# Wet weight over time
phenotypes %>%
  group_by(temperature) %>%
  dplyr::summarise(sl_mean.1=mean(sl_11212022),sl_sd.1=sd(sl_11212022), 
            sl_mean.2=mean(sl_12272022),sl_sd.2=sd(sl_12272022),
            sl_mean.3=mean(sl_final),sl_sd.3=sd(sl_final),
            
            wwt_mean.1=mean(wwt_11212022),wwt_sd.1=sd(wwt_11212022), 
            wwt_mean.2=mean(wwt_12272022),wwt_sd.2=sd(wwt_12272022),
            wwt_mean.3=mean(wwt_final),wwt_sd.3=sd(wwt_final)) %>%
  pivot_longer(cols = -temperature) %>%
  separate(name, sep = "\\.", into = c("metric", "time")) %>%
  mutate_at(c("metric"), factor) %>% mutate(time=as.numeric(time)) %>%
  filter(metric=="wwt_mean") %>%
  ggplot() + geom_line(aes(x=time, y=value, color=temperature), cex=1.5) + theme_minimal() +
  scale_color_manual(name="Temperature", 
                       values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) + 
  ggtitle("Wet weight over time") + xlab(NULL) + ylab("Wet weight") +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("Pre-Acclimation", "Pre-Treatment", "Treatment\ntermination"))
```

```{r, message=F, warning=F}
### Standard length by temperature over time - mean and SD
phenotypes %>%
  group_by(temperature) %>%
  dplyr::summarise(sl_mean.1=mean(sl_11212022),sl_sd.1=sd(sl_11212022), 
            sl_mean.2=mean(sl_12272022),sl_sd.2=sd(sl_12272022),
            sl_mean.3=mean(sl_final),sl_sd.3=sd(sl_final),
            
            wwt_mean.1=mean(wwt_11212022),wwt_sd.1=sd(wwt_11212022), 
            wwt_mean.2=mean(wwt_12272022),wwt_sd.2=sd(wwt_12272022),
            wwt_mean.3=mean(wwt_final),wwt_sd.3=sd(wwt_final)) %>%
  pivot_longer(cols = -temperature) %>%
  separate(name, sep = "\\.", into = c("metric", "time")) %>%
  mutate_at(c("metric"), factor) %>% mutate(time=as.numeric(time)) %>%
  filter(metric=="sl_mean") %>%
  ggplot() + geom_line(aes(x=time, y=value, color=temperature), cex=1.5) + theme_minimal() +
  scale_color_manual(name="Temperature", 
                       values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) + 
  ggtitle("Standard length over time") + xlab(NULL) + ylab("Standard length") +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("Pre-Acclimation", "Pre-Treatment", "Treatment\ntermination"))
```

#### Specific growth while acclimating, Standard Length

Statistics - did it differ among treatments? No. 

```{r, message=F, warning=F}
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(SGR.sl.accl), sd=sd(SGR.sl.accl))
summary(aov(SGR.sl.accl ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(SGR.sl.accl ~ temperature, phenotypes %>% filter(mort=="No")))

```

```{r, message=F, warning=F}
phenotypes %>% 
    ggplot(aes(y=SGR.sl.accl, x=temperature, fill=temperature, shape=mort, alpha=mort)) + 
  geom_violin(trim = F) + geom_point(position = position_jitterdodge(jitter.width = 1)) + theme_minimal() +
    ggtitle("Specific growth rate while acclimating, standard length (~ %/day)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_x_discrete(drop=T) + #Do drop empty factors for temperature contrasts
  ylab("mm / day") +
  scale_fill_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) + ylab(NULL)

# phenotypes %>% filter(mort=="No") %>%
#     ggplot(aes(y=SGR.sl.accl/days_growth, x=temperature, fill=temperature)) + 
#   geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
#     ggtitle("Growth rate while acclimating, standard length (mm/day)") +
#   scale_fill_manual(name="Temperature", 
#                        values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) + 
#   xlab(NULL) + ylab(NULL)
```
#### Specific growth in treatments, Standard Length

Did SGR(sl) differ among treatments? Yes! Among all pairwise comparisons. 

Note: should see if this data fits the model/estimates using this formula from Laurel et al. 2015, https://doi.org/10.1007/s00300-015-1761-5 
G indiv = y o  + aT + bT 2 + cT 3

```{r, message=F, warning=F}
summary(aov(SGR.sl.trt ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(SGR.sl.trt ~ temperature, phenotypes %>% filter(mort=="No")))
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(SGR.sl.trt), sd=sd(SGR.sl.trt))
```

```{r, message=F, warning=F}
phenotypes %>% 
    ggplot(aes(y=SGR.sl.trt, x=temperature, fill=temperature, shape=mort, alpha=mort)) + 
  geom_violin() + geom_point(position = position_jitterdodge(jitter.width = 1)) + theme_minimal() +
    ggtitle("Specific growth rate in treatments, standard length (~%day)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  ylab("mm / day") +
  scale_fill_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) + ylab(NULL)
```
```{r}
SGR.sl.model <- lm(SGR.sl.trt ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes)
summary(SGR.sl.model)

SGR.sl.predict <- data.frame(temperature=seq(0,16,length.out=100))
SGR.sl.predict$predicted <- predict(SGR.sl.model, SGR.sl.predict)

phenotypes %>% 
    ggplot(aes(y=SGR.sl.trt, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Specific growth rate in treatments, standard length (~%day)") +
  geom_line(data=SGR.sl.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("% / day")
```
```{r}
round(SGR.sl.model$coefficients, digits=3) %>% as.vector() %>% as.data.frame(row.names = c("y","a","b","c")) %>% set_names(NULL) %>% kable(caption="Model coefficients")

# What is the maximum standard length growth rate and associated predicted temperature as per this model?  0.35%/day at 12C.
SGR.sl.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted growth rate & associated temperature, standard length, %/day")
```

#### Specific growth while acclimating, wet weight

No difference among temperatures 

```{r, message=F, warning=F}
summary(aov(SGR.ww.accl ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(SGR.ww.accl ~ temperature, phenotypes %>% filter(mort=="No")))
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(SGR.ww.accl), sd=sd(SGR.ww.accl))
```

```{r}
phenotypes %>% 
    ggplot(aes(y=SGR.ww.accl, x=temperature, fill=temperature, shape=mort, alpha=mort)) + 
  geom_violin() + geom_point(position = position_jitterdodge(jitter.width = 1)) + theme_minimal() +
    ggtitle("Growth rate while acclimating, wet weight (g/day)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  ylab("g / day") +
  scale_fill_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) 
```
#### Specific growth in treatments, Wet Weight

Did SGR(wet weight) differ among treatments? Yes! Among all pairwise comparisons. 

Note: should see if this data fits the model/estimates using this formula from Laurel et al. 2015, https://doi.org/10.1007/s00300-015-1761-5 
G indiv = y o  + aT + bT 2 + cT 3

```{r, message=F, warning=F}
summary(aov(SGR.ww.trt ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(SGR.ww.trt ~ temperature, phenotypes %>% filter(mort=="No")))
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(SGR.ww.trt), sd=sd(SGR.ww.trt))
```

```{r, message=F, warning=F}
phenotypes %>% 
    ggplot(aes(y=SGR.ww.trt, x=temperature, fill=temperature, shape=mort, alpha=mort)) + 
  geom_violin() + geom_point(position = position_jitterdodge(jitter.width = 1)) + theme_minimal() +
    ggtitle("Specific growth rate in treatments, wet weight (~%/day)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  ylab("mm / day") +
  scale_fill_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) + ylab(NULL)
```
```{r}
SGR.ww.model <- lm(SGR.ww.trt ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes)
summary(SGR.ww.model)

SGR.ww.predict <- data.frame(temperature=seq(0,16,length.out=100))
SGR.ww.predict$predicted <- predict(SGR.ww.model, SGR.ww.predict)

phenotypes %>% 
    ggplot(aes(y=SGR.ww.trt, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Specific growth rate in treatments, wet weight (~%/day)") +
  geom_line(data=SGR.ww.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("% / day")
```
```{r}
round(SGR.ww.model$coefficients, digits=3) %>% as.vector() %>% as.data.frame(row.names = c("y","a","b","c")) %>% set_names(NULL) %>% kable(caption="Model coefficients")

# What is the maximum standard length growth rate and associated predicted temperature as per this model?  0.35%/day at 12C.
SGR.ww.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted growth rate & associated temperature, wet weight, %/day")
```

#### Combined growth rate plot 

```{r}
phenotypes %>% 
  group_by(temperature) %>% summarise(
    SGR.sl.mean=mean(SGR.sl.trt), SGR.sl.sd=sd(SGR.sl.trt),
    SGR.ww.mean=mean(SGR.ww.trt), SGR.ww.sd=sd(SGR.ww.trt)) %>%  
  ggplot() + 
  geom_point(aes(x=as.numeric(as.character(temperature)), y=SGR.sl.mean), size=3) + 
  geom_point(aes(x=as.numeric(as.character(temperature)), y=SGR.ww.mean), size=3) + 
  geom_errorbar(aes(x = as.numeric(as.character(temperature)), ymin = SGR.sl.mean - SGR.sl.sd, ymax = SGR.sl.mean + SGR.sl.sd), width = 0.2) +
  geom_errorbar(aes(x = as.numeric(as.character(temperature)), ymin = SGR.ww.mean - SGR.ww.sd, ymax = SGR.ww.mean + SGR.ww.sd), width = 0.2) +
  theme_minimal() +
  ggtitle("Specific growth rate in treatments, ~% per day") +
  geom_line(data=SGR.sl.predict, aes(x=temperature, y=predicted), color="gray25", size=1) +
  geom_line(data=SGR.ww.predict, aes(x=temperature, y=predicted), color="gray25", size=1) +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_x_continuous(breaks = c(seq(0,16,2))) +
  xlab("Temperature") +   ylab("% / day") + 
  geom_text(aes(x = 13.6, y = 0.83, label="Standard Length"), size=4, color="gray25") +
  geom_text(aes(x = 13.2, y = 0.39, label="Wet Weight"), size=4, color="gray25") +
  geom_vline(xintercept = SGR.sl.predict[which.max(SGR.sl.predict$predicted),]$temperature, 
             linetype = "dashed", color = "gray25", size = 1) +
  geom_vline(xintercept = SGR.ww.predict[which.max(SGR.ww.predict$predicted),]$temperature, 
             linetype = "dashed", color = "gray25", size = 1)
```


#### Body condition index at termination (Kwet)

Not significantly affected by temperature (but marginal, p=0.07)

```{r, message=F, warning=F}
summary(aov(Kwet ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(Kwet ~ temperature, phenotypes %>% filter(mort=="No")))
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(Kwet), sd=sd(Kwet))
```

```{r, message=F, warning=F}
phenotypes %>% 
    ggplot(aes(y=Kwet, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Body condition index\n(wet weight / standard length") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
       values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("Wet weight / standard length")
```
<!-- Model Kwet by temperature -->

<!-- ```{r} -->
<!-- K.model <- lm(Kwet ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2) -->
<!-- summary(K.model) -->

<!-- K.predict <- data.frame(temperature=seq(0,16,length.out=100)) -->
<!-- K.predict$predicted <- predict(K.model, K.predict) -->

<!-- # What is the maximum standard length growth rate and associated predicted temperature as per this model?  0.35%/day at 12C. -->
<!-- K.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted composite performance index & associated temperature, wet weight, %/day") -->

<!-- phenotypes2 %>%  -->
<!--     ggplot(aes(y=Kwet, x=as.numeric(as.character(temperature)))) +  -->
<!--   geom_point(aes(color=temperature)) + theme_minimal() + -->
<!--   ggtitle("Kwet (condition index)") + -->
<!--   geom_line(data=K.predict, aes(x=temperature, y=predicted), color="gray45", size=1) + -->
<!--   #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") + -->
<!--   scale_alpha_manual(values=c(0.75,0.5)) + -->
<!--   scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) + -->
<!--   xlab(NULL) +   ylab("% / day") -->
<!-- ``` -->

#### Hepato-somatic index 

Very different among temperature treatments except for 0 vs. 5C. 

```{r, message=F, warning=F}
summary(aov(hsi ~ temperature, phenotypes %>% filter(mort=="No")))
TukeyHSD(aov(hsi ~ temperature, phenotypes %>% filter(mort=="No")))
phenotypes %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(hsi), sd=sd(hsi))
```

```{r, message=F, warning=F}
phenotypes %>%
    ggplot(aes(y=hsi, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Hepato-somatic index\n(100 * liver wet weight / total wet weight)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
       values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("100 * Liver wet weight / Total wet weight")

# phenotypes %>% filter(mort=="No") %>%
#     ggplot(aes(y=hsi, x=temperature, fill=temperature)) + 
#   geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
#     ggtitle("Appr. hepato-somatic index\n(liver wet weight / total wet weight)") +
#   scale_fill_manual(name="Temperature", 
#                        values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c"))
```
<!-- Model HSI by temperature -->

<!-- ```{r} -->
<!-- HSI.model <- lm(hsi ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2) -->
<!-- summary(HSI.model) -->

<!-- HSI.predict <- data.frame(temperature=seq(0,16,length.out=100)) -->
<!-- HSI.predict$predicted <- predict(HSI.model, HSI.predict) -->

<!-- # What is the maximum standard length growth rate and associated predicted temperature as per this model?  0.35%/day at 12C. -->
<!-- HSI.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted composite performance index & associated temperature, wet weight, %/day") -->

<!-- phenotypes2 %>%  -->
<!--     ggplot(aes(y=hsi, x=as.numeric(as.character(temperature)))) +  -->
<!--   geom_point(aes(color=temperature)) + theme_minimal() + -->
<!--   ggtitle("HSI (liver condition index)") + -->
<!--   geom_line(data=HSI.predict, aes(x=temperature, y=predicted), color="gray45", size=1) + -->
<!--   #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") + -->
<!--   scale_alpha_manual(values=c(0.75,0.5)) + -->
<!--   scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) + -->
<!--   xlab(NULL) +   ylab("% / day") -->
<!-- ``` -->

### Lipid data! 

```{r}
lipids <- read_excel("../data/Lipid class liver data_091324.xlsx") %>% clean_names() %>%
  mutate(sample_id=as.factor(sample_id))
```

#### Add lipid data to phenotype data, and calculate composite performance index 

Composite Performance Index calculated as SGR(length) * Kwet * HSI * 

```{r}
phenotypes2 <- phenotypes %>% 
  left_join(lipids %>% dplyr::select(sample_id, genetics_sample_number, tag_20:total_32), 
            by=c("microchip_id"="sample_id")) %>%
    mutate(G.sl.scaled=rescale(SGR.sl.trt, to=c(0,1)),
           G.ww.scaled=rescale(SGR.ww.trt, to=c(0,1)),
           Kwet.scaled=rescale(Kwet, to=c(0,1)),
           hsi.scaled=rescale(hsi, to=c(0,1)),
           total.lipid.scaled=rescale(total_32, to=c(0,1))) %>% 
    mutate(pi=100*G.sl.scaled*G.ww.scaled*Kwet.scaled*hsi.scaled*total.lipid.scaled)

head(phenotypes2)
```

### Liver lipids consitutents 

#### Total Lipids

```{r}
summary(aov(total_32 ~ temperature, phenotypes2 %>% filter(mort=="No")))
TukeyHSD(aov(total_32 ~ temperature, phenotypes2 %>% filter(mort=="No")))
phenotypes2 %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(total_32, na.rm=T), sd=sd(total_32, na.rm=T))

phenotypes2 %>% 
  filter(!is.na(total_32)) %>%
    ggplot(aes(y=total_32, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Total lipid content (ug/mg)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("Total lipid Content")

total.model <- lm(total_32 ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2)
summary(total.model)
total.predict <- data.frame(temperature=seq(0,16,length.out=100))
total.predict$predicted <- predict(total.model, total.predict)

# What is the maximum total lipid and associated predicted temperature as per this model?  0.35%/day at 12C.
total.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted total lipid & associated temperature")

phenotypes2 %>% 
    ggplot(aes(y=total_32, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Total lipid content") +
  geom_line(data=total.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("% / day")
```

#### Triglycerides

```{r}
summary(aov(tag_28 ~ temperature, phenotypes2 %>% filter(mort=="No")))
TukeyHSD(aov(tag_28 ~ temperature, phenotypes2 %>% filter(mort=="No")))
phenotypes2 %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(tag_28, na.rm=T), sd=sd(tag_28, na.rm=T))

phenotypes2 %>% 
  filter(!is.na(total_32)) %>%
    ggplot(aes(y=tag_28, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Triglyceride content (ug/mg)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("TAG Content")

tag.model <- lm(tag_28 ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2)
summary(tag.model)
tag.predict <- data.frame(temperature=seq(0,16,length.out=100))
tag.predict$predicted <- predict(tag.model, tag.predict)

# What is the maximum triglyceride and associated predicted temperature as per this model?  0.35%/day at 12C.
tag.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted triglyceride & associated temperature")

phenotypes2 %>% 
    ggplot(aes(y=tag_28, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Triglyceride content") +
  geom_line(data=tag.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("TAG Content")
```

#### Free Fatty Acids

```{r}
summary(aov(ffa_29 ~ temperature, phenotypes2 %>% filter(mort=="No")))
TukeyHSD(aov(Kwet ~ temperature, phenotypes2 %>% filter(mort=="No")))
phenotypes2 %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(total_32, na.rm=T), sd=sd(total_32, na.rm=T))

phenotypes2 %>% 
  filter(!is.na(total_32)) %>%
    ggplot(aes(y=ffa_29, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Free fatty acid content (ug/mg)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("FFA Content")

ffa.model <- lm(ffa_29 ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2)
summary(ffa.model)
ffa.predict <- data.frame(temperature=seq(0,16,length.out=100))
ffa.predict$predicted <- predict(ffa.model, ffa.predict)

# What is the maximum FFA and associated predicted temperature as per this model?  0.35%/day at 12C.
ffa.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted free fatty acid & associated temperature")

phenotypes2 %>% 
    ggplot(aes(y=ffa_29, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Free fatty acid content") +
  geom_line(data=ffa.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("FFA Content")
```

#### Sterols 

```{r}
summary(aov(st_30 ~ temperature, phenotypes2 %>% filter(mort=="No")))
TukeyHSD(aov(st_30 ~ temperature, phenotypes2 %>% filter(mort=="No")))
phenotypes2 %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(st_30, na.rm=T), sd=sd(st_30, na.rm=T))

phenotypes2 %>% 
  filter(!is.na(total_32)) %>%
    ggplot(aes(y=st_30, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Sterol content (ug/mg)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("sterol Content")

st.model <- lm(st_30 ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2)
summary(st.model)
st.predict <- data.frame(temperature=seq(0,16,length.out=100))
st.predict$predicted <- predict(st.model, st.predict)

# What is the maximum sterol and associated predicted temperature as per this model?  0.35%/day at 12C.
st.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted sterol & associated temperature")

phenotypes2 %>% 
    ggplot(aes(y=st_30, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Sterol content") +
  geom_line(data=st.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("Sterol Content")
```

#### Polar Lipids 

```{r}
summary(aov(polar_lipids_31 ~ temperature, phenotypes2 %>% filter(mort=="No")))
TukeyHSD(aov(polar_lipids_31 ~ temperature, phenotypes2 %>% filter(mort=="No")))
phenotypes2 %>% filter(mort=="No") %>% group_by(temperature) %>% dplyr::summarise(mean=mean(total_32, na.rm=T), sd=sd(total_32, na.rm=T))

phenotypes2 %>% 
  filter(!is.na(total_32)) %>%
    ggplot(aes(y=polar_lipids_31, x=interaction(mort, temperature), fill=temperature, shape=mort, alpha=mort, label=genetic_sampling_count)) + 
  geom_violin() + geom_jitter(width=0.25) + theme_minimal() +
    ggtitle("Polar lipid content (ug/mg)") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab(NULL) + ylab("Polar lipid Content")
```

#### Composite Performance Index 

```{r, warning=F, message=F}
phenotypes2 %>% ggplot(aes(x=temperature, y=pi, fill=temperature)) + 
  geom_violin(alpha=0.6) + geom_jitter(width = 0.2) + 
    scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme_minimal() + ylab("Composite Performance Index\nGSR(sl) * GSR(ww) * Kwet * HSI * Total Lipid")
```

Model composite performance index by temperature

```{r, warning=F, message=F}
pi.model <- lm(pi ~ poly(as.numeric(as.character(temperature)), 3), data = phenotypes2)
summary(pi.model)

pi.predict <- data.frame(temperature=seq(0,16,length.out=100))
pi.predict$predicted <- predict(pi.model, pi.predict)

# What is the maximum standard length growth rate and associated predicted temperature as per this model?  0.35%/day at 12C.
pi.predict %>% arrange(desc(predicted)) %>% head(n=1) %>% round(digits = 2) %>% kable(caption = "Maximum predicted composite performance index & associated temperature, wet weight, %/day")

phenotypes2 %>% 
    ggplot(aes(y=pi, x=as.numeric(as.character(temperature)))) + 
  geom_point(aes(color=temperature)) + theme_minimal() +
  ggtitle("Composite Performance Index") +
  geom_line(data=pi.predict, aes(x=temperature, y=predicted), color="gray45", size=1) +
  #  stat_smooth(method="lm", se=TRUE, formula=y ~ poly(x, 3),colour="gray50") +
  scale_alpha_manual(values=c(0.75,0.5)) +
  scale_color_manual(name="Temperature", values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  xlab(NULL) +   ylab("% / day")
```


```{r, warning=F, message=F}
phenotypes2 %>% dplyr::select(microchip_id, genetic_sampling_count, temperature, G.sl.scaled:pi) %>% 
  pivot_longer(cols = c(G.sl.scaled:total.lipid.scaled), names_to = "biometric", values_to = "scaled_value") %>% 
  ggplot(aes(y=pi, x=scaled_value, color=temperature)) + geom_point() + 
  facet_wrap(~biometric) + 
  theme_minimal_grid() + xlab("Scaled value") + ylab("Composite Performancde Index") + 
    scale_color_manual(name="Temperature", 
        values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  geom_smooth(method = "glm", se = F) + theme(legend.position = "bottomright")
```

```{r}
phenotypes3 <- phenotypes2 %>% select(microchip_id, genetic_sampling_count, temperature, SGR.sl.trt, SGR.ww.trt, Kwet, hsi, mort, tag_28, ffa_29, st_30, polar_lipids_31, total_32, pi) %>%
  filter(mort=="No")
save(phenotypes3, file="../data/phenotypes3")
```

### All biometrics in one plot 

```{r, warning=FALSE, message=FALSE}
phenotypes3 %>% filter(mort=="No") %>%  dplyr::select(-mort) %>% 
  dplyr::rename("K(wet)"="Kwet", "HSI"="hsi", "Triglycerides"="tag_28", "Free Fatty Acids"="ffa_29", 
         "Sterols"="st_30", "Polar Lipids"="polar_lipids_31", "Total Lipids"="total_32",
         "Performance Index"="pi", "Growth Rate, Length"="SGR.sl.trt",
         "Growth Rate, Weight"="SGR.ww.trt") %>%  
  select(-`Performance Index`) %>% 
  pivot_longer(`Growth Rate, Length`:HSI, names_to = "metric", values_to = "value") %>% 
  mutate(metric=as.factor(metric)) %>%
 
  ggplot(aes(x=temperature, y=value, fill=temperature)) + 
  #geom_jitter(width=0.25, size=1.5) + 
  geom_boxplot(alpha=0.5) + 
  theme_minimal() +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom") +
  facet_wrap(~metric, scales = "free", nrow=3) + xlab(NULL) + ylab(NULL)
```
### All lipid metrics in one plot 

```{r, warning=FALSE, message=FALSE}
phenotypes3 %>% filter(mort=="No") %>%  dplyr::select(-mort) %>% 
  dplyr::rename("K(wet)"="Kwet", "HSI"="hsi", "Triglycerides"="tag_28", "Free Fatty Acids"="ffa_29", 
         "Sterols"="st_30", "Polar Lipids"="polar_lipids_31", "Total Lipids"="total_32",
         "Performance Index"="pi", "Growth Rate, Length"="SGR.sl.trt",
         "Growth Rate, Weight"="SGR.ww.trt") %>%  
  select(-`Performance Index`) %>% 
  pivot_longer(`Triglycerides`:`Total Lipids`, names_to = "metric", values_to = "value") %>% 
  mutate(metric=as.factor(metric)) %>%
 
  ggplot(aes(x=temperature, y=value, fill=temperature)) + 
  #geom_jitter(width=0.25, size=1.5) + 
  geom_boxplot(alpha=0.5) + 
  theme_minimal() +
  scale_fill_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom") +
  facet_wrap(~metric, scales = "free", nrow=3) + xlab(NULL) + ylab(NULL)
```
### Correlations among metrics 

```{r, message=F, warning=F}
require(corrplot)
corr.metrics <- cor(phenotypes2 %>% select(SGR.sl.trt, SGR.ww.trt, Kwet, hsi, mort, days_growth, tag_28, ffa_29, st_30, polar_lipids_31, total_32, pi) %>%
  filter(mort=="No") %>% mutate(growthrate.sl=SGR.sl.trt/days_growth, growthrate.wwt=SGR.ww.trt/days_growth) %>% select(-SGR.sl.trt, -SGR.ww.trt, -mort, -days_growth), use="complete.obs")

corrplot::corrplot(corr.metrics, method = "number", type="lower", diag=T, tl.cex=.8, number.cex = 0.8, sig.level = 0.05, tl.pos="tl")
corrplot::corrplot(corr.metrics, method = "color", type="upper", diag=F, sig.level = 0.05, add=T, tl.pos="n", )
```

### Residuals, all metrics ~ temperature 

```{r}
a <- lm(total_32~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
b <- lm(polar_lipids_31~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
c <- lm(st_30~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
d <- lm(ffa_29~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
e <- lm(tag_28~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
f <- lm(Kwet~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
g <- lm(hsi~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
h <- lm(`SGR.sl.trt`~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))
i <- lm(`SGR.ww.trt`~as.numeric(as.character(temperature)), data=phenotypes3 %>% filter(!is.na(total_32)))

phenotypes3.resids <- cbind(
  sample=as.character((phenotypes3 %>% filter(!is.na(total_32)))$microchip_id),
  temperature=as.character((phenotypes3 %>% filter(!is.na(total_32)))$temperature),        
        resid_total.lipids=resid(a),
         resid_polar=resid(b),
         resid_sterol=resid(c),
         resid_ffa=resid(d),
         resid_trigl=resid(e),
         resid_Kwet=resid(f),
         resid_HSI=resid(g),
         resid_sl.growthrate=resid(h),
         resid_wwt.growthrate=resid(i)) %>% as.data.frame()
```

```{r, warning=F, message=F}
phenotypes3.resids %>% 
  mutate(temperature=factor(temperature, ordered=T, levels=c(0,5,9,16))) %>%
  pivot_longer(starts_with("resid"), names_to = "metric", values_to = "resid") %>% 
  mutate(temperature=as.factor(temperature), metric=as.factor(metric), resid=as.numeric(resid)) %>% 
 
  ggplot(aes(x=temperature, y=resid, color=temperature, text=sample)) + 
  geom_jitter(width=0.25, size=1.5) + theme_minimal() +
  scale_color_manual(name="Temperature", 
     values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  facet_wrap(~metric, scales = "free")
```

### Correlations, residuals ~ temperature 

```{r}
corr.metrics.resids <- cor(phenotypes3.resids %>% select(-sample, -temperature) %>% mutate_all(as.numeric), use="complete.obs")

corrplot::corrplot(corr.metrics.resids, method = "number", type="lower", diag=T, tl.cex=.8, number.cex = 0.8, sig.level = 0.05, tl.pos="tl")
corrplot::corrplot(corr.metrics.resids, method = "color", type="upper", diag=F, sig.level = 0.05, add = TRUE, tl.pos = "n")
```

```{r}
# anova(a) #total lipid
# anova(b) #polar lipids <-- only metric not affected by temperature 
# anova(c) #sterols 
# anova(d) #free fatty acids 
# anova(e) #triglycerols 
# anova(f) #condition index
# anova(g) #HSI
# anova(h) #growth rate - length
# anova(i) #growth rate - weight
```

#### Look at total lipids ~ condition index  

```{r, warning=F, message=F}
#ggplotly(
  phenotypes3 %>% 
    ggplot(aes(y=total_32, x=Kwet, color=temperature)) + 
  geom_point() + theme_minimal() +
    ggtitle("Total Lipids / Condition Index") +
  scale_color_manual(name="Temperature", 
        values=c("0"="#2c7bb6", "5"="#abd9e9", "9"="#fdae61", "16"="#d7191c")) +
  geom_smooth(method = "lm") +   facet_wrap(~temperature) +
  xlab("Condition Index") + ylab("Total Lipids")#)

```
